### 2024-06-18 y1review.tex
add initial draft text

\section{Introduction}

Formal verification is crucial in ensuring the reliability and
correctness of software systems. In the context of compiler design, this
importance is magnified as compilers are responsible for translating
high-level code into machine-executable instructions. Reasoning about
compiler optimizations is notoriously difficult, and even minor bugs in
widely used compilers can lead to subtle, hard-to-debug miscompilations
with far-reaching consequences. This challenge makes the formal
verification of compiler intermediate representations (IRs) an
especially compelling topic, as it promises to enhance the dependability
of the entire software development lifecycle.

Now, formal verification has been done before. Most notable is the
CompCert project: a fully verified end-to-end C compiler.
{[}{[}@leroyCompCertFormallyVerified{]}{]} Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

Take, for example, LLVM, a widely-used open-source compiler that is used
to compile, e.g., C, C++, Fortran, Rust, and Swift. Verifying the entire
LLVM codebase, comprising of a few million lines of C++ code, is
infeasible. Instead, we follow the same approach as Lopez et al and
focus on verifying that individual rewrites on a compiler IR preserve
semantics. {[}{[}@lopesAlive2BoundedTranslation2021{]}{]}

Although LLVM is indeed used by the standard compilers for Rust and
Swift, both also use other, language-specific, IRs to be able to do
analyses and optimizations at a higher level of abstraction. In
response, the MLIR (Multi-Level Intermediate Representation\footnote{Multi-Level
  Intermediate Representation is the \emph{official} expansion of MLIR,
  but other expansions are used by the community as well}) project has
been created as a framework for creating domain-specific IRs (generally
called a \emph{dialect}) at various levels of abstraction.

During the first year of my PhD studies, my research has focused on the
LeanMLIR project, a framework for defining semantics for MLIR dialects
within the Lean theorem prover. A central aspect of this work has been
modelling MLIR dialects with completely pure semantics, exploiting the
higher-level nature of MLIR dialect, where referential transparency is
encouraged.

We need the full power of a theorem prover to be able to capture the
full extensible nature of MLIR. Still, to recover the ease-of-use that
SMT-solver-based tools provide, we've also been investing in building
solid automation for bitvector proofs in collaboration with the Lean
community.

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are statefull, and cycles are broken by looking at the value of
a variable at a previous state. For example, graph regions are used by
dialects that model hardware {[}{[}@eldridgeMLIRHardwareCompiler{]}{]},
where variables model wires and cycles represent physical cycles in the
circuit.

Since the exact semantics of graph regions may differ between dialects,
we need some general framework to reason about stateful behaviours in
Lean. I propose that coinductive types are this framework, and would
like to continue my earlier work on supporting coinductive reasoning in
Lean, with the aim of using them to model graph regions in LeanMLIR.

In fact, coinductive types are useful beyond graph regions. Since
coinduction models infinite behaviours, it is also a good way to model
non-termination without giving up on computability of a dialect's
semantics, thus addressing a current short-coming of LeanMLIR.

\hypertarget{related-work}{%
\section{Related work}\label{related-work}}

First, we'll review the literature on verified compilation, which
naturally leads us to the literature on coinductive types.

\hypertarget{verified-compilation}{%
\subsection{Verified Compilation}\label{verified-compilation}}

{[}{[}@leroyCompCertFormallyVerified\textbar CompCert{]}{]} is the
seminal work when it comes to verified compilation: an end-to-end
verified to not miscompile compiler for the C language. However, exactly
because it's an end-to-end verified monolith, CompCert is not
competitive compared to traditional non-verified compilers. Outside of
certain high-risk domains, where traditionally optimizations would be
turned off entirely {[}{[}@kastnerCompCertPracticalExperience{]}{]}, the
vast majority of C code is thus still compiled with non-verified
compilers, which studies have shown to be prone to miscompilations
{[}{[}@yangFindingUnderstandingBugs2011{]}{]}.

Lopes et al.~propose \emph{translation validation} as a way to scale
formal verification. {[}{[}@lopesAlive2BoundedTranslation2021{]}{]}
Their implementation, called Alive, has found wide adoption in the
development of LLVM. Given a declarative description of an optimization
as a source and target pattern in LLVM IR, the tool uses an SMT solver
to automatically determine whether the optimized pattern exhibits only
behaviours of the original program.

Alive is able to leverate SMT solvers because the semantics of LLVM IR
are given in terms of bitvectors, and SMT solvers understand the
bitvector domain well. However, the semantics of more high-level
compiler IRs, for which MLIR was created, can involve complex
mathematical objects, which today's SMT solvers are not able to prove
properties about. For example, the semantics for the ``Poly'' dialect
\footnote{https://homomorphicencryption.org/} are generally given in
terms of finitely-presented commutative rings. Furthermore, MLIR is, by
design, extensible, so a translation validation tool for MLIR also needs
to be easily extensible to cover new dialects.

Our group has been developing
{[}\protect\hyperlink{leanmlir}{LeanMLIR}{]}, a framework for giving
semantics to MLIR-style compiler IRs in the Lean theorem prover
{[}{[}@demouraLeanTheoremProver2015{]}{]}. As a theorem prover, Lean is
much more expressible than SMT solvers. Among theorem provers, Lean is
particularly suited for this purpose because of it's community of
mathematicians developing a comprehensive mathematical library
{[}{[}@themathlibcommunityLeanMathematicalLibrary2020{]}{]}, and because
of it's highly extensive syntax and meta-programming capabilities
{[}{[}@ullrichNotationsHygienicMacro2022{]}{]}{[}{[}@paulinoMetaprogrammingLean{]}{]}.

However, using a theorem prover means we don't get as much automation
out of the box, and one area where Lean is still lacking is in its lack
of support for coinductive types.

\hypertarget{coinductive-types}{%
\subsection{Coinductive Types}\label{coinductive-types}}

Lean, like most theorem provers, is built with around a native concept
of inductive types. This is no surprise, given that most mathematical
constructions are inductive in nature. Coinductive constructions, in
contrast, are less common in general mathematics, but, they are crucial
in computer science.

For us, the main difference of interest is that inductive datastructures
are finite, while coinductive datastructures capture infinite
behaviours.
{[}{[}@sangiorgiIntroductionBisimulationCoinduction2011{]}{]} In
particular, since Lean requires us to prove that recursive functions are
terminating, we need corecursion to correctly model things like infinite
loops and other non-terminating behaviours that are common in compiler
IRs.

For example, Xia et al.~used interaction trees, a coinductive type, in
their Coq formalization of recursive and impure programs.
{[}{[}@xiaInteractionTreesRepresenting2020{]}{]} Similarly, MLIR has an
``fsm'' dialect \footnote{https://circt.llvm.org/docs/Dialects/FSM/}
centred on finite state machines. Such automata are understood to be
coinductive in nature, so mechanizing the semanantics of this dialect in
Lean will likely require coinduction.

Consequently, we need a user-friendly way to define coinductive
structures and to formalize coinductive proof methods in Lean. Key prior
work in this area include my own master's thesis
{[}{[}@keizerImplementingDefinitionalCo{]}{]}, which is based on earlier
work by Avigad et al that models coinductive as quotients of polynomial
functors (QPFs){[}{[}@avigadDataTypesQuotients2019{]}{]}.

Crucially, this prior work builds coinductive types using constructions
already supported by Lean today. This allows us to provide coinductive
types as a \emph{library}, instead of having to modify the logical
system itself to support coinduction natively --- an approach that is
greatly aided by Lean's excellent meta-programming capabilities
mentioned previously.

When looking at coinductive types in other theorem provers, we
distinguish between languages like Isabelle
{[}{[}@traytelCategoryTheoryBased{]}{]}, which follow the same
coinduction-as-a-library approach, and languages like Coq
{[}{[}@gimenezTutorialRecursiveTypes1998{]}{]}{[}{[}@gimenezApplicationCoinductiveTypes1996{]}{]}
or Agda which have modified their trusted kernels to support
coinduction.

Isabelle, in particular, is relevant because it's construction of
coinductive types, in terms of \emph{bounded natural functors} is
closely related to our construction in terms of QPFs.
{[}{[}@furerQuotientsBoundedNatural2022{]}{]} There are still new
challenges, however, because Lean's dependent types mean we have to
worry about the distinction between definitional equality (which is a
decidable, but weaker notion of equality) and propositional equality
(which captures more things that are intuitively equal, but requires
proof). Isabelle, in contrast, uses a weaker logical system where all
equalities are decidable.

In contrast, the coinduction-in-the-kernel approach allows us to modify
the kernel to understand coinduction and ensure the kernel recognizes
the desired definitonal equalities. However, modifying the trusted
kernel carries a large burden of proof, since the changes could
compromise the logical soundness of the entire system. In fact, the
original implementation of coinduction, called \emph{positive}
coinduction, is nowadays discouraged in favour of an alternative
implementation, since positive coinduction breaks the subject reduction
property \footnote{ The subject reduction property states that reducing
  a program does not change its type}{[}{[}@sozeauCorrectCompleteType{]}{]}.
Our coinductives-as-a-library approach requires no new axioms nor
changes to the kernel, and is thus guaranteed not to change any meta
properties of Lean.

\hypertarget{progress-report}{%
\section{Progress Report}\label{progress-report}}

The bulk of my time has been spent on developing, in collaboration with
other members of our research group, the LeanMLIR framework, culminating
in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
Compiler IRs''}. As part of this development, we

\hypertarget{lean-mlir}{%
\subsection{Lean MLIR}\label{lean-mlir}}

LeanMLIR is a framework for giving semantics to MLIR dialects in Lean.
On top of this framework, we've built a peephole rewriter, and proved
that if a rewrite locally preserves semantics, then applying this
rewrite in the context of a larger program will preserve the semantics
of the program as a whole.

However, we've focused on dialects with pure (i.e., side-effect free)
semantics. Even though we \emph{are} able to verify a lot of interesting
rewrites in high-level MLIR dialects, we do need to be able to reason
about side-effect if we want to capture the low-level dialects as well,
as a reviewer of the paper rightly pointed out.

To this end, we've generalised the framework so that a dialect's
semantics may be given in a user-defined monad, and proven that applying
a semantics-preserving pure rewrite (i.e., where the operations being
rewritten are all side-effect free) inside of a side-effecting program
preserves semantics of the program as a whole. However, this
generalisation still has a lot of rough edges, and further work is
needed before we can claim our framework is versatile enough to deal
with side-effects cleanly.

\hypertarget{bitvectors}{%
\subsection{Bitvectors}\label{bitvectors}}

As mentioned, a big reason for us to use Lean is it's community. As
such, I've been heavily involved in the redesign of Lean's API for
bitvectors, and the move of bitvectors from Mathlib (the mathematical
library previously mentioned) to become part of Lean's core library
(i.e., the library that is by default included in every Lean program).
By ensuring our representation of bitvectors is aligned with the rest of
the community, we ensure we can easily reap the benefits of, e.g.,
tactics to automate bitvector proofs made by others.

One such automation effort is the LeanSAT\footnote{https://github.com/leanprover/leansat}
library, which aims to automate proofs about bitvectors of bounded
widths, by implementing a verified bitblaster and SAT sovler in Lean.
Siddharth (another PhD student in our group) and I have been
contributing to this project by formally proving necessary facts in the
core library.

This automation is crucial, because with it we can provide push-button
automation for non-expert users on the same level as Alive, for dialects
whose semantics use bitvectors, while also still having the full
expressivity needed for dialects that use more complex mathematical
objects. Thus, addressing the main drawback of using a theorem prover.

\hypertarget{future-plans}{%
\section{Future plans}\label{future-plans}}

\hypertarget{bitvectors-1}{%
\subsection{Bitvectors}\label{bitvectors-1}}

\begin{quote}
June '24 - January '25
\end{quote}

As mentioned in the progress report, I'm currently contributing to a
verified bitblasting procedure in Lean, called LeanSAT. Once this is
implemented, I plan to write a paper contrasting this verified
bitblaster and SAT solver with popular non-verified SMT solvers like Z3,
to investigate what exactly the cost is of the extra assurance that a
verified implementation gives. This is particularly relevant, since,
just like compilers, SMT solvers are highly complicated pieces of
software in which bugs are known to exist
{[}{[}@brummayerFuzzingDeltadebuggingSMT2009{]}{]}.

\hypertarget{arbitrary-widths}{%
\subsubsection{Arbitrary widths}\label{arbitrary-widths}}

LeanSAT is, however, not a silver bullet. Just like SMT solvers, it can
only reason about bitvector expressions where the width of the
bitvectors involved is bounded. One of the topics I'd like to explore
more is the connection between bitvectors and 2-adic numbers (which are
infinite and thus coinductive in nature!) to build a decision procedure
which can decide equality of a certain (limited) class of
arbitrary-width bitvector expressions.

Beyond being more satisfying, this arbitrary-width procedure is likely
more performant than SAT solving, and thus relevant even if the end-user
isn't necessarily interested in proving their property for all possible
widths.

\hypertarget{amazon-internship}{%
\subsubsection{Amazon Internship}\label{amazon-internship}}

\begin{quote}
July '24 - November '24
\end{quote}

Furthermore, I plan to continue this work on bitvector automation in
Lean during my internship at Amazon \footnote{ Assuming I'll get
  permission to work away, of course}. They are particularly interested
in automatically reasoning about ARM assembly code, which closely aligns
with our goal of automating bitvector reasoning to then use for
verifying (low-level) MLIR dialects.

\hypertarget{coinduction}{%
\subsection{Coinduction}\label{coinduction}}

The next big theme to tackle will be to make the Lean library for
coinductive types ready for serious use, which mainly involves
high-level support for corecursive functions.

\hypertarget{productivity-checking}{%
\subsubsection{Productivity Checking}\label{productivity-checking}}

\begin{quote}
June '24 - Aug '24 (Supervising Internship)
\end{quote}

Currently, the \texttt{QpfTypes} framework is able to compile a
high-level description of a coinductive type into a construction in
terms of concepts Lean already supports. As part of this construction, a
\emph{corecursion principle} is generated. For example, the corecursion
principle for \texttt{Stream\ \textbackslash{}alpha}, the type of
infinite lists (i.e., streams) over elements of type
\texttt{\textbackslash{}alpha}, is as follows:

\begin{verbatim}
Stream.corec : (\beta -> \alpha \times \beta) -> \beta -> Stream \a
\end{verbatim}

While this corecursion principle is in theory all that's needed to
define functions over streams, it's not directly usable in practice.
Hence, the first subgoal will be to support a high-level way to define
corecursive functions, following the idiomatic way of defining inductive
functions in terms of pattern matching, rather than recursion
principles, which is then compiled into an application of this
corecursion princple. This will involve some software engineering,
dealing with Lean's extensible syntax, plus a good amount of research,
to know which functions are \emph{productive} in our formalization, and
to automatically recognize (a well-defined sub-class of) these
productive corecursive functions.

Productivity in this context is the dual of termination in recursive
contexts; a corecursive function is said to be productive if, at all
times, it will produce the next element of our infinite corecursive
structure in finite time. Ensuring all corecursive functions are
productive is paramount to our logical system remaining consistent. In
fact, because of our coinductives-as-a-library approach we are unable to
define corecursive functions without proving they are productive, and
conversely, if we can get the system to accept a corecursive function,
we know that allowing this function must be productive, and allowing it
is sound.

I will be a co-supervising an undergraduate student over the summer who
will be implementing support for corecursive functions using a
\emph{guarded} productivity checker. That is, this first implementation
will only allow only functions where each corecursive call occurs behind
a constructor. This check is known to incomplete, and more complicated
schemes have been studied
({[}{[}@blanchetteFriendsBenefitsImplementing2017{]}{]}), but a
guardedness easier to implement and will most likely be sufficient for
now --- the initial implementation of corecursion in Coq also allowed
only guarded corecursion {[}{[}@gimenezTutorialRecursiveTypes1998{]}{]}.
Of course, if there is still time left in the internship after this goal
has been reached, we will attempt to implement more complicated schemes
as well.

\hypertarget{runtime-behaviour}{%
\subsubsection{Runtime behaviour}\label{runtime-behaviour}}

\begin{quote}
Jan '25 - May '25
\end{quote}

A cornerstone of the coinduction-as-a-library approach is that we're
limited to whatever concepts our theorem prover already supports. The
QPF construction starts by realizing that an element of an inductive
type is basically a (finite) tree, while an element of a coinductive
type is a non-wellfounded (i.e., possibly infinite) tree.

Finite trees are easily encoded in Lean, but representing a finite tree
is a bit more tricky. The current construction works by representing an
non-wellfounded tree as a function from a natural number \(n\) to the
approximation of the non-wellfounded tree up to a maximal depth of
\(n\), together with a proof that each subsequent approximation agrees
with the previous approximations up to depth \(n-1\).

This construction is mathematically elegant, but seems to lead to
unexpected inefficient computational behaviour, as shown by some early
experimentation by myself. I plan to investigate this more thoroughly,
but I strongly suspect this behaviour comes from the impedance mismatch
between the construction of infinite trees as just described, and the
fact that applications of corecursion principles are essentially state
machines. I would thus like to explore the possibility of building the
QPF constructions on a representation of non-wellfounded trees in terms
of state machines (or, more generally, coalgebras).

\hypertarget{leanmlir}{%
\subsection{LeanMLIR}\label{leanmlir}}

Finally, we return to the main objective: formalizing compiler IRs.

\hypertarget{non-termination}{%
\subsubsection{Non-termination}\label{non-termination}}

\begin{quote}
Apr '25 - June '25
\end{quote}

In the status report, we discussed how the framework was generalized to
work in arbitrary monads. But, to actually define the semantics of an
MLIR dialect which allows for infinite loops, we need a concrete monad
that can support this behaviour.

We could, of course, model non-termination using the \texttt{Option}
monad, where divergence is simply modelled as \texttt{Option.none}. Such
semantics are necessarily non-computable, losing the very nice property
that the formalization of a dialect in LeanMLIR also gives a reference
implementation for an interpreter.

Therefore, I'd like to use my library for coinductive types to define a
coinductive monad which supports computable infinite behaviours, drawing
inspiration from the interaction tree library for coq.
{[}{[}@xiaInteractionTreesRepresenting2020{]}{]} This addresses a
weakness in LeanMLIR while simultaneously giving a nice case study for
the usability of the coinduction library.

\hypertarget{graph-regions-parametric-coinduction}{%
\subsubsection{Graph Regions \& Parametric
Coinduction}\label{graph-regions-parametric-coinduction}}

\begin{quote}
June '25 - Sep '25
\end{quote}

A \emph{Graph Region} in MLIR is, just like a regular region, a sequence
of instructions with their results bound to variables. However, unlike
normal regions, in a graph region the \emph{use} of a variable may
precede it's definition. For example

\begin{verbatim}
^bb(%x : i32)
  %y = comb.and %x %y : (i32, i32) -> (i32)
\end{verbatim}

Here, \texttt{y} is obtained as the bitwise and of the argument
\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
sense, but this style of definition is common in hardware
representations, where this is interpreted as: ``the value of \texttt{y}
in the \emph{current} clock cycle is computed by taking the bitwise and
of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''
\footnote{This example is simplified, in reality you'd have to specify a
  register instruction to control when a value is stored across clock
  cycles. See https://circt.llvm.org/ for more details}. To formally
model this, we could denote variables as infinite streams of bitvectors,
which gives us the value at every possible future clock cycle. Another
example of a dialect with graph regions is the FSM, or Finite State
Machine, dialect.

Graph regions are thus inherently stateful, and modelling them is a
particularly good use-case for coinductive types and coinductive
reasoning. To facilitate easy verification of coinductive properties in
this setting, I would like to investigate supporting parametrized
coinductive proofs {[}{[}@hurPowerParameterizationCoinductive{]}{]} in
the coinductive types framework.

\hypertarget{stretch-goal-iris}{%
\subsection{Stretch Goal: Iris}\label{stretch-goal-iris}}

\begin{quote}
{[}!TODO{]} Talk about how denotational semantics are cool, and how the
aforementioned improvements will let us define dialects with
side-effects, but reasoning about them will still be a bit painful. We'd
really like to reason with program logics instead, in particular, iris:
{[}{[}@jungIrisMonoidsInvariants2015{]}{]}{[}{[}@jungHigherorderGhostState2016{]}{]}
\end{quote}

### 2024-06-18 paper.tex
fix citations

--- 2024-06-18 y1review.tex
@@
+\def\paperversiondraft{draft}
+\def\paperversionnormal{normal}
+
+% If the paper version is set to 'normal' mode keep it,
+% otherwise set it to 'draft' mode.
+\ifx\paperversion\paperversionnormal
+\else
+  \def\paperversion{draft}
+\fi
+
+\documentclass[a4paper]{scrartcl}
+
+\usepackage{colortbl}
+
+% 'draftonly' environment
+\usepackage{environ}
+\ifx\paperversion\paperversiondraft
+\newenvironment{draftonly}{}{}
+\else
+\NewEnviron{draftonly}{}
+\fi
+
+% Most PL conferences are edited by conference-publishing.com. Follow their
+% advice to add the following packages.
+%
+% The first enables the use of UTF-8 as character encoding, which is the
+% standard nowadays. The second ensures the use of font encodings that support
+% accented characters etc. (Why should I use this?). The mictotype package
+% enables certain features 'to­wards ty­po­graph­i­cal per­fec­tion
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{microtype}
+
+\usepackage{xargs}
+\usepackage{lipsum}
+\usepackage{xparse}
+\usepackage{xifthen, xstring}
+\usepackage{xspace}
+\usepackage{marginnote}
+\usepackage{etoolbox}
+\usepackage[acronym,shortcuts]{glossaries}
+
+\input{tex/setup.tex}
+% \input{tex/acm.tex}
+
+\usemintedstyle{colorful}
+
+% Newer versions of minted require the 'customlexer' argument for custom lexers
+% whereas older versions require the '-x' to be passed via the command line.
+\makeatletter
+\ifcsdef{minted@optlistcl@quote}
+{
+\newminted[mlir]{tools/MLIRLexer.py:MLIRLexerOnlyOps}{customlexer, mathescape}
+\newminted[xdsl]{tools/MLIRLexer.py:MLIRLexer}{customlexer, mathescape, style=murphy}
+\newminted[lean4]{tools/Lean4Lexer.py:Lean4Lexer}{customlexer, mathescape}
+}
+{
+\newminted[mlir]{tools/MLIRLexer.py:MLIRLexerOnlyOps -x}{mathescape}
+\newminted[xdsl]{tools/MLIRLexer.py:MLIRLexer -x}{mathescape, style=murphy}
+\newminted[lean4]{tools/Lean4Lexer.py:Lean4Lexer -x}{mathescape}
+}
+\makeatother
+
+% We use the following color scheme
+% 
+% This scheme is both print-friendly and colorblind safe for
+% up to four colors (including the red tones makes it not
+% colorblind safe any more)
+%
+% https://colorbrewer2.org/#type=qualitative&scheme=Paired&n=4
+
+\definecolor{pairedNegOneLightGray}{HTML}{cacaca}
+\definecolor{pairedNegTwoDarkGray}{HTML}{827b7b}
+\definecolor{pairedOneLightBlue}{HTML}{a6cee3}
+\definecolor{pairedTwoDarkBlue}{HTML}{1f78b4}
+\definecolor{pairedThreeLightGreen}{HTML}{b2df8a}
+\definecolor{pairedFourDarkGreen}{HTML}{33a02c}
+\definecolor{pairedFiveLightRed}{HTML}{fb9a99}
+\definecolor{pairedSixDarkRed}{HTML}{e31a1c}
+
+\createtodoauthor{grosser}{pairedOneLightBlue}
+\createtodoauthor{authorTwo}{pairedTwoDarkBlue}
+\createtodoauthor{authorThree}{pairedThreeLightGreen}
+\createtodoauthor{authorFour}{pairedFourDarkGreen}
+\createtodoauthor{authorFive}{pairedFiveLightRed}
+\createtodoauthor{authorSix}{pairedSixDarkRed}
+
+\newacronym{ir}{IR}{Intermediate Representation}
+
+\graphicspath{{./images/}}
+
+% Define macros that are used in this paper
+%
+% We require all macros to end with a delimiter (by default {}) to enusure
+% that LaTeX adds whitespace correctly.
+\makeatletter
+\newcommand\requiredelimiter[2][########]{%
+  \ifdefined#2%
+    \def\@temp{\def#2#1}%
+    \expandafter\@temp\expandafter{#2}%
+  \else
+    \@latex@error{\noexpand#2undefined}\@ehc
+  \fi
+}
+\@onlypreamble\requiredelimiter
+\makeatother
+
+\newcommand\newdelimitedcommand[2]{
+\expandafter\newcommand\csname #1\endcsname{#2}
+\expandafter\requiredelimiter
+\csname #1 \endcsname
+}
+
+\newdelimitedcommand{leanmlir}{LeanMLIR}
+
+\usepackage{booktabs}
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+
+\usepackage[verbose]{newunicodechar}
+\newunicodechar{₁}{\ensuremath{_1}}
+\newunicodechar{₂}{\ensuremath{_2}}
+\newunicodechar{∀}{\ensuremath{\forall}}
+\newunicodechar{α}{\ensuremath{\alpha}}
+\newunicodechar{β}{\ensuremath{\beta}}
+
+% \circled command to print a colored circle.
+% \circled{1} pretty-prints "(1)"
+% This is useful to refer to labels that are embedded within figures.
+\DeclareRobustCommand{\circled}[2][]{%
+    \ifthenelse{\isempty{#1}}%
+        {\circledbase{pairedOneLightBlue}{#2}}%
+        {\autoref{#1}: \hyperref[#1]{\circledbase{pairedOneLightBlue}{#2}}}%
+}
+
+% listings don't write "Listing" in autoref without this.
+\providecommand*{\listingautorefname}{Listing}
+\newcommand{\sectionautorefname}{Section}
+\newcommand{\subsectionautorefname}{Section}
+\newcommand{\subsubsectionautorefname}{Section}
+
+\begin{document}
+
+%% Title information
+% \title{First Year Report}       %% [Short Title] is optional;
+                                      %% when present, will be used in
+                                      %% header instead of Full Title.
+\title{Towards a framework for reasoning about stateful compiler IRs 
+using coinductive reasoning.}                                      
+\subtitle{First Year Report}
+
+\author{Alex Keizer \small{(ack55)}}
+\date{}
+
+% \begin{abstract}
+% % An abstract should consist of six main sentences:
+% %  1. Introduction. In one sentence, what’s the topic?
+% %  2. State the problem you tackle.
+% %  3. Summarize (in one sentence) why nobody else has adequately answered the research question yet.
+% %  4. Explain, in one sentence, how you tackled the research question.
+% %  5. In one sentence, how did you go about doing the research that follows from your big idea.
+% %  6. As a single sentence, what’s the key impact of your research?
+
+% % (http://www.easterbrook.ca/steve/2010/01/how-to-write-a-scientific-abstract-in-six-easy-steps/)
+
+% \end{abstract}
+\maketitle
+
 \section{Introduction}
@@
 Now, formal verification has been done before. Most notable is the
-CompCert project: a fully verified end-to-end C compiler.
-{[}{[}@leroyCompCertFormallyVerified{]}{]} Yet, the vast majority of
+CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
 code today is still being compiled with non-verified compilers. We ask,
@@
 focus on verifying that individual rewrites on a compiler IR preserve
-semantics. {[}{[}@lopesAlive2BoundedTranslation2021{]}{]}
+semantics.~\cite{lopesAlive2BoundedTranslation2021}
 
@@
 a variable at a previous state. For example, graph regions are used by
-dialects that model hardware {[}{[}@eldridgeMLIRHardwareCompiler{]}{]},
+dialects that model hardware,
 where variables model wires and cycles represent physical cycles in the
-circuit.
+circuit.~\cite{eldridgeMLIRHardwareCompiler}
 
@@
 
-\hypertarget{related-work}{%
-\section{Related work}\label{related-work}}
+\section{Related work}\label{related-work}
 
@@
 
-\hypertarget{verified-compilation}{%
-\subsection{Verified Compilation}\label{verified-compilation}}
+\subsection{Verified Compilation}\label{verified-compilation}
 
-{[}{[}@leroyCompCertFormallyVerified\textbar CompCert{]}{]} is the
+CompCert is the
 seminal work when it comes to verified compilation: an end-to-end
-verified to not miscompile compiler for the C language. However, exactly
+verified to not miscompile compiler for the C language. \cite{leroyCompCertFormallyVerified}
+However, exactly
 because it's an end-to-end verified monolith, CompCert is not
@@
 certain high-risk domains, where traditionally optimizations would be
-turned off entirely {[}{[}@kastnerCompCertPracticalExperience{]}{]}, the
+turned off entirely \cite{kastnerCompCertPracticalExperience}, the
 vast majority of C code is thus still compiled with non-verified
 compilers, which studies have shown to be prone to miscompilations
-{[}{[}@yangFindingUnderstandingBugs2011{]}{]}.
+\cite{yangFindingUnderstandingBugs2011}.
 
 Lopes et al.~propose \emph{translation validation} as a way to scale
-formal verification. {[}{[}@lopesAlive2BoundedTranslation2021{]}{]}
+formal verification. \cite{lopesAlive2BoundedTranslation2021}
 Their implementation, called Alive, has found wide adoption in the
@@
 Our group has been developing
-{[}\protect\hyperlink{leanmlir}{LeanMLIR}{]}, a framework for giving
-semantics to MLIR-style compiler IRs in the Lean theorem prover
-{[}{[}@demouraLeanTheoremProver2015{]}{]}. As a theorem prover, Lean is
+LeanMLIR, a framework for giving
+semantics to MLIR-style compiler IRs in the Lean theorem prover.~\cite{demouraLeanTheoremProver2015} As a theorem prover, Lean is
 much more expressible than SMT solvers. Among theorem provers, Lean is
 particularly suited for this purpose because of it's community of
-mathematicians developing a comprehensive mathematical library
-{[}{[}@themathlibcommunityLeanMathematicalLibrary2020{]}{]}, and because
-of it's highly extensive syntax and meta-programming capabilities
-{[}{[}@ullrichNotationsHygienicMacro2022{]}{]}{[}{[}@paulinoMetaprogrammingLean{]}{]}.
+mathematicians developing a comprehensive mathematical library,~\cite{themathlibcommunityLeanMathematicalLibrary2020} 
+and because of it's highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.
 
@@
 
-\hypertarget{coinductive-types}{%
-\subsection{Coinductive Types}\label{coinductive-types}}
+\subsection{Coinductive Types}\label{coinductive-types}
 
@@
 behaviours.
-{[}{[}@sangiorgiIntroductionBisimulationCoinduction2011{]}{]} In
+\cite{sangiorgiIntroductionBisimulationCoinduction2011} In
 particular, since Lean requires us to prove that recursive functions are
@@
 their Coq formalization of recursive and impure programs.
-{[}{[}@xiaInteractionTreesRepresenting2020{]}{]} Similarly, MLIR has an
+\cite{xiaInteractionTreesRepresenting2020} Similarly, MLIR has an
 ``fsm'' dialect \footnote{https://circt.llvm.org/docs/Dialects/FSM/}
@@
 work in this area include my own master's thesis
-{[}{[}@keizerImplementingDefinitionalCo{]}{]}, which is based on earlier
+\cite{keizerImplementingDefinitionalCo}, which is based on earlier
 work by Avigad et al that models coinductive as quotients of polynomial
-functors (QPFs){[}{[}@avigadDataTypesQuotients2019{]}{]}.
+functors (QPFs)\cite{avigadDataTypesQuotients2019}.
 
@@
 distinguish between languages like Isabelle
-{[}{[}@traytelCategoryTheoryBased{]}{]}, which follow the same
+\cite{traytelCategoryTheoryBased}, which follow the same
 coinduction-as-a-library approach, and languages like Coq
-{[}{[}@gimenezTutorialRecursiveTypes1998{]}{]}{[}{[}@gimenezApplicationCoinductiveTypes1996{]}{]}
+\cite{gimenezTutorialRecursiveTypes1998}\cite{gimenezApplicationCoinductiveTypes1996}
 or Agda which have modified their trusted kernels to support
@@
 closely related to our construction in terms of QPFs.
-{[}{[}@furerQuotientsBoundedNatural2022{]}{]} There are still new
+\cite{furerQuotientsBoundedNatural2022} There are still new
 challenges, however, because Lean's dependent types mean we have to
@@
 property \footnote{ The subject reduction property states that reducing
-  a program does not change its type}{[}{[}@sozeauCorrectCompleteType{]}{]}.
+  a program does not change its type}.~\cite{sozeauCorrectCompleteType}
 Our coinductives-as-a-library approach requires no new axioms nor
@@
 
-\hypertarget{progress-report}{%
-\section{Progress Report}\label{progress-report}}
+\section{Progress Report}\label{progress-report}
 
@@
 
-\hypertarget{lean-mlir}{%
-\subsection{Lean MLIR}\label{lean-mlir}}
+\subsection{Lean MLIR}\label{lean-mlir}
 
@@
 
-\hypertarget{bitvectors}{%
-\subsection{Bitvectors}\label{bitvectors}}
+\subsection{Bitvectors}\label{bitvectors}
 
@@
 
-\hypertarget{future-plans}{%
-\section{Future plans}\label{future-plans}}
+\section{Future plans}\label{future-plans}
 
-\hypertarget{bitvectors-1}{%
-\subsection{Bitvectors}\label{bitvectors-1}}
+\subsection{Bitvectors}\label{bitvectors-1}
 
@@
 software in which bugs are known to exist
-{[}{[}@brummayerFuzzingDeltadebuggingSMT2009{]}{]}.
+\cite{brummayerFuzzingDeltadebuggingSMT2009}.
 
-\hypertarget{arbitrary-widths}{%
-\subsubsection{Arbitrary widths}\label{arbitrary-widths}}
+\subsubsection{Arbitrary widths}\label{arbitrary-widths}
 
@@
 
-\hypertarget{amazon-internship}{%
-\subsubsection{Amazon Internship}\label{amazon-internship}}
+\subsubsection{Amazon Internship}\label{amazon-internship}
 
@@
 
-\hypertarget{coinduction}{%
-\subsection{Coinduction}\label{coinduction}}
+\subsection{Coinduction}\label{coinduction}
 
@@
 
-\hypertarget{productivity-checking}{%
-\subsubsection{Productivity Checking}\label{productivity-checking}}
+\subsubsection{Productivity Checking}\label{productivity-checking}
 
@@
 schemes have been studied
-({[}{[}@blanchetteFriendsBenefitsImplementing2017{]}{]}), but a
+(\cite{blanchetteFriendsBenefitsImplementing2017}), but a
 guardedness easier to implement and will most likely be sufficient for
 now --- the initial implementation of corecursion in Coq also allowed
-only guarded corecursion {[}{[}@gimenezTutorialRecursiveTypes1998{]}{]}.
+only guarded corecursion \cite{gimenezTutorialRecursiveTypes1998}.
 Of course, if there is still time left in the internship after this goal
@@
 
-\hypertarget{runtime-behaviour}{%
-\subsubsection{Runtime behaviour}\label{runtime-behaviour}}
-
+\subsubsection{Runtime behaviour}\label{runtime-behaviour}
 \begin{quote}
@@
 
-\hypertarget{leanmlir}{%
-\subsection{LeanMLIR}\label{leanmlir}}
+\subsection{LeanMLIR}\label{leanmlir}
 
@@
 
-\hypertarget{non-termination}{%
-\subsubsection{Non-termination}\label{non-termination}}
+\subsubsection{Non-termination}\label{non-termination}
 
@@
 inspiration from the interaction tree library for coq.
-{[}{[}@xiaInteractionTreesRepresenting2020{]}{]} This addresses a
+\cite{xiaInteractionTreesRepresenting2020} This addresses a
 weakness in LeanMLIR while simultaneously giving a nice case study for
@@
 
-\hypertarget{graph-regions-parametric-coinduction}{%
 \subsubsection{Graph Regions \& Parametric
-Coinduction}\label{graph-regions-parametric-coinduction}}
+Coinduction}\label{graph-regions-parametric-coinduction}
 
@@
 this setting, I would like to investigate supporting parametrized
-coinductive proofs {[}{[}@hurPowerParameterizationCoinductive{]}{]} in
+coinductive proofs~\cite{hurPowerParameterizationCoinductive} in
 the coinductive types framework.
 
-\hypertarget{stretch-goal-iris}{%
-\subsection{Stretch Goal: Iris}\label{stretch-goal-iris}}
+\subsection{Stretch Goal: Iris}\label{stretch-goal-iris}
 
@@
 really like to reason with program logics instead, in particular, iris:
-{[}{[}@jungIrisMonoidsInvariants2015{]}{]}{[}{[}@jungHigherorderGhostState2016{]}{]}
+\cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
 \end{quote}
+
+% \input{y1review.tex}
+
+% \begin{figure}
+% % Link to figure
+% %
+% % https://docs.google.com/drawings/d/1juKp43D3rLC-luBQPwQZ_wCnDK2S_6C1k6USV0wKE0g/edit?usp=sharing
+% \includegraphics[width=\columnwidth]{overview.pdf}
+% \caption{Our key idea visualized}
+% \grosser{Replace this figure with your own drawing.}
+% \end{figure}
+
+%% Bibliography
+\bibliographystyle{plain}
+\bibliography{references}
+
+\end{document}

### 2024-06-18 paper.tex
fix citations for real
use bibTex format instead of bibLaTex

@@
 \usepackage[acronym,shortcuts]{glossaries}
+\usepackage{hyperref}
 
@@
 \providecommand*{\listingautorefname}{Listing}
-\newcommand{\sectionautorefname}{Section}
-\newcommand{\subsectionautorefname}{Section}
-\newcommand{\subsubsectionautorefname}{Section}
+\renewcommand{\sectionautorefname}{Section}
+\renewcommand{\subsectionautorefname}{Section}
+\renewcommand{\subsubsectionautorefname}{Section}
 
@@
 in the \emph{current} clock cycle is computed by taking the bitwise and
-of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''
-\footnote{This example is simplified, in reality you'd have to specify a
+of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{
+  This example is simplified, in reality you'd have to specify a
   register instruction to control when a value is stored across clock
-  cycles. See https://circt.llvm.org/ for more details}. To formally
-model this, we could denote variables as infinite streams of bitvectors,
+  cycles. See https://circt.llvm.org/ for more details}
+To formally model this, we could denote variables as infinite streams of bitvectors,
 which gives us the value at every possible future clock cycle. Another

### 2024-06-18 paper.tex
whitespace

@@
 seminal work when it comes to verified compilation: an end-to-end
-verified to not miscompile compiler for the C language. \cite{leroyCompCertFormallyVerified}
+verified to not miscompile compiler for the C language.~\cite{leroyCompCertFormallyVerified}
 However, exactly
@@
 certain high-risk domains, where traditionally optimizations would be
-turned off entirely \cite{kastnerCompCertPracticalExperience}, the
+turned off entirely,~\cite{kastnerCompCertPracticalExperience} the
 vast majority of C code is thus still compiled with non-verified
-compilers, which studies have shown to be prone to miscompilations
-\cite{yangFindingUnderstandingBugs2011}.
+compilers, which studies have shown to be prone to miscompilations.~\cite{yangFindingUnderstandingBugs2011}
 
 Lopes et al.~propose \emph{translation validation} as a way to scale
-formal verification. \cite{lopesAlive2BoundedTranslation2021}
+formal verification.~\cite{lopesAlive2BoundedTranslation2021}
 Their implementation, called Alive, has found wide adoption in the
@@
 are finite, while coinductive datastructures capture infinite
-behaviours.
-\cite{sangiorgiIntroductionBisimulationCoinduction2011} In
-particular, since Lean requires us to prove that recursive functions are
+behaviours.~\cite{sangiorgiIntroductionBisimulationCoinduction2011}
+In particular, since Lean requires us to prove that recursive functions are
 terminating, we need corecursion to correctly model things like infinite
@@
 For example, Xia et al.~used interaction trees, a coinductive type, in
-their Coq formalization of recursive and impure programs.
-\cite{xiaInteractionTreesRepresenting2020} Similarly, MLIR has an
-``fsm'' dialect \footnote{https://circt.llvm.org/docs/Dialects/FSM/}
+their Coq formalization of recursive and impure programs.~\cite{xiaInteractionTreesRepresenting2020} Similarly, MLIR has an
+``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
 centred on finite state machines. Such automata are understood to be
@@
 structures and to formalize coinductive proof methods in Lean. Key prior
-work in this area include my own master's thesis
-\cite{keizerImplementingDefinitionalCo}, which is based on earlier
+work in this area include my own master's thesis (\cite{keizerImplementingDefinitionalCo}), 
+which is based on earlier
 work by Avigad et al that models coinductive as quotients of polynomial
-functors (QPFs)\cite{avigadDataTypesQuotients2019}.
+functors (QPFs).~\cite{avigadDataTypesQuotients2019}
 
@@
 When looking at coinductive types in other theorem provers, we
-distinguish between languages like Isabelle
-\cite{traytelCategoryTheoryBased}, which follow the same
-coinduction-as-a-library approach, and languages like Coq
-\cite{gimenezTutorialRecursiveTypes1998}\cite{gimenezApplicationCoinductiveTypes1996}
+distinguish between languages like Isabelle~\cite{traytelCategoryTheoryBased}, which follow the same
+coinduction-as-a-library approach, and languages like Coq~\cite{gimenezTutorialRecursiveTypes1998,gimenezApplicationCoinductiveTypes1996}
 or Agda which have modified their trusted kernels to support
@@
 coinductive types, in terms of \emph{bounded natural functors} is
-closely related to our construction in terms of QPFs.
-\cite{furerQuotientsBoundedNatural2022} There are still new
+closely related to our construction in terms of QPFs.~\cite{furerQuotientsBoundedNatural2022} 
+There are still new
 challenges, however, because Lean's dependent types mean we have to
@@
 implementation, since positive coinduction breaks the subject reduction
-property \footnote{ The subject reduction property states that reducing
-  a program does not change its type}.~\cite{sozeauCorrectCompleteType}
+property\footnote{The subject reduction property states that reducing a program does not change its type}.~\cite{sozeauCorrectCompleteType}
 Our coinductives-as-a-library approach requires no new axioms nor

### 2024-06-25 paper.tex
put references before the dot spelling and grammar pass comment out iris

@@
 \usepackage{hyperref}
+
+% \usepackage[numbers]{natbib} % for \citeauthor
 
@@
 
+\newcommand*{\etal}{~\emph{et al.}}
+
 \begin{document}
@@
 Formal verification is crucial in ensuring the reliability and
-correctness of software systems. In the context of compiler design, this
-importance is magnified as compilers are responsible for translating
-high-level code into machine-executable instructions. Reasoning about
-compiler optimizations is notoriously difficult, and even minor bugs in
-widely used compilers can lead to subtle, hard-to-debug miscompilations
-with far-reaching consequences. This challenge makes the formal
-verification of compiler intermediate representations (IRs) an
-especially compelling topic, as it promises to enhance the dependability
-of the entire software development lifecycle.
+correctness of software systems. 
+Reasoning about compiler optimizations, in particular, is notoriously difficult, 
+and even minor bugs in widely-used compilers can lead to subtle, hard-to-debug miscompilations
+with far-reaching consequences. 
+This challenge makes the formal verification of compiler intermediate representations (IRs) a
+particularly compelling topic, 
+as it promises to enhance the dependability of the entire software development lifecycle.
 
 Now, formal verification has been done before. Most notable is the
-CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
+CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}. Yet, the vast majority of
 code today is still being compiled with non-verified compilers. We ask,
-thus, how do we formally verify not \emph{a} compiler, but, how do we
+thus, how do we formally verify not just \emph{a} compiler, but, how do we
 scale formal verification to \emph{mainstream} compilers.
@@
 LLVM codebase, comprising of a few million lines of C++ code, is
-infeasible. Instead, we follow the same approach as Lopez et al and
-focus on verifying that individual rewrites on a compiler IR preserve
-semantics.~\cite{lopesAlive2BoundedTranslation2021}
+infeasible. Instead, we follow the same approach as Lopes\etal{} by
+focussing on verifying that individual rewrites on a compiler IR preserve
+semantics~\cite{lopesAlive2BoundedTranslation2021}.
 
@@
 where variables model wires and cycles represent physical cycles in the
-circuit.~\cite{eldridgeMLIRHardwareCompiler}
+circuit~\cite{eldridgeMLIRHardwareCompiler}.
 
@@
 seminal work when it comes to verified compilation: an end-to-end
-verified to not miscompile compiler for the C language.~\cite{leroyCompCertFormallyVerified}
+verified to not miscompile compiler for the C language~\cite{leroyCompCertFormallyVerified}.
 However, exactly
@@
 vast majority of C code is thus still compiled with non-verified
-compilers, which studies have shown to be prone to miscompilations.~\cite{yangFindingUnderstandingBugs2011}
+compilers, which studies have shown to be prone to miscompilations~\cite{yangFindingUnderstandingBugs2011}.
 
 Lopes et al.~propose \emph{translation validation} as a way to scale
-formal verification.~\cite{lopesAlive2BoundedTranslation2021}
+formal verification~\cite{lopesAlive2BoundedTranslation2021}.
 Their implementation, called Alive, has found wide adoption in the
@@
 LeanMLIR, a framework for giving
-semantics to MLIR-style compiler IRs in the Lean theorem prover.~\cite{demouraLeanTheoremProver2015} As a theorem prover, Lean is
+semantics to MLIR-style compiler IRs in the Lean theorem prover~\cite{demouraLeanTheoremProver2015}. 
+As a theorem prover, Lean is
 much more expressible than SMT solvers. Among theorem provers, Lean is
@@
 mathematicians developing a comprehensive mathematical library,~\cite{themathlibcommunityLeanMathematicalLibrary2020} 
-and because of it's highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.
+and because of its highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.
 
@@
 
-Lean, like most theorem provers, is built with around a native concept
+Lean, like most theorem provers, is built around a native concept
 of inductive types. This is no surprise, given that most mathematical
@@
 are finite, while coinductive datastructures capture infinite
-behaviours.~\cite{sangiorgiIntroductionBisimulationCoinduction2011}
+behaviours~\cite{sangiorgiIntroductionBisimulationCoinduction2011}.
 In particular, since Lean requires us to prove that recursive functions are
@@
 For example, Xia et al.~used interaction trees, a coinductive type, in
-their Coq formalization of recursive and impure programs.~\cite{xiaInteractionTreesRepresenting2020} Similarly, MLIR has an
+their Coq formalization of recursive and impure programs~\cite{xiaInteractionTreesRepresenting2020}.
+Similarly, MLIR has an
 ``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
@@
 structures and to formalize coinductive proof methods in Lean. Key prior
-work in this area include my own master's thesis (\cite{keizerImplementingDefinitionalCo}), 
+work in this area includes my own master's thesis (\cite{keizerImplementingDefinitionalCo}), 
 which is based on earlier
 work by Avigad et al that models coinductive as quotients of polynomial
-functors (QPFs).~\cite{avigadDataTypesQuotients2019}
+functors (QPFs)~\cite{avigadDataTypesQuotients2019}.
 
@@
 coinductive types, in terms of \emph{bounded natural functors} is
-closely related to our construction in terms of QPFs.~\cite{furerQuotientsBoundedNatural2022} 
+closely related to our construction in terms of QPFs~\cite{furerQuotientsBoundedNatural2022}. 
 There are still new
@@
 implementation, since positive coinduction breaks the subject reduction
-property\footnote{The subject reduction property states that reducing a program does not change its type}.~\cite{sozeauCorrectCompleteType}
+property\footnote{The subject reduction property states that reducing a program does not change its type}~\cite{sozeauCorrectCompleteType}.
 Our coinductives-as-a-library approach requires no new axioms nor
@@
 
-The bulk of my time has been spent on developing, in collaboration with
-other members of our research group, the LeanMLIR framework, culminating
-in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
-Compiler IRs''}. As part of this development, we
+The bulk of my time has been spent developing the LeanMLIR framework, 
+in collaboration with other members of our research group, 
+culminating in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
+Compiler IRs''}.
 
@@
 
-\subsection{Stretch Goal: Iris}\label{stretch-goal-iris}
+% \subsection{Stretch Goal: Iris}\label{stretch-goal-iris}
 
-\begin{quote}
-{[}!TODO{]} Talk about how denotational semantics are cool, and how the
-aforementioned improvements will let us define dialects with
-side-effects, but reasoning about them will still be a bit painful. We'd
-really like to reason with program logics instead, in particular, iris:
-\cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
-\end{quote}
+% \begin{quote}
+% {[}!TODO{]} Talk about how denotational semantics are cool, and how the
+% aforementioned improvements will let us define dialects with
+% side-effects, but reasoning about them will still be a bit painful. We'd
+% really like to reason with program logics instead, in particular, iris:
+% \cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
+% \end{quote}
 

### 2024-06-25 paper.tex
shuffle projects around

--- 2024-06-18 paper.tex
@@
 
-Now, formal verification has been done before. Most notable is the
-CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
+Now, formally verified compilation has been done before.
+Most notable is the CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
 code today is still being compiled with non-verified compilers. We ask,
@@
 We need the full power of a theorem prover to be able to capture the
-full extensible nature of MLIR. Still, to recover the ease-of-use that
-SMT-solver-based tools provide, we've also been investing in building
-solid automation for bitvector proofs in collaboration with the Lean
-community.
+extensible and high-level nature of MLIR dialects.
+Still, to recover the ease-of-use that SMT-solver-based tools provide, 
+we've also been investing in building solid automation for bitvector proofs in collaboration with the Lean community.
 
@@
 def-use chain into a general graph. Generally, the semantics of such
-dialects are statefull, and cycles are broken by looking at the value of
-a variable at a previous state. For example, graph regions are used by
+dialects are stateful, and cycles are broken by looking at the value of
+a variable at a previous state.
+For example, graph regions are used by
 dialects that model hardware,
@@
 in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
-Compiler IRs''}. As part of this development, we
+Compiler IRs''}.
 
@@
 \begin{quote}
-June '24 - January '25
+June '24 - November '24
 \end{quote}
@@
 As mentioned in the progress report, I'm currently contributing to a
-verified bitblasting procedure in Lean, called LeanSAT. Once this is
-implemented, I plan to write a paper contrasting this verified
-bitblaster and SAT solver with popular non-verified SMT solvers like Z3,
-to investigate what exactly the cost is of the extra assurance that a
-verified implementation gives. This is particularly relevant, since,
-just like compilers, SMT solvers are highly complicated pieces of
-software in which bugs are known to exist
-\cite{brummayerFuzzingDeltadebuggingSMT2009}.
+verified bitblasting procedure in Lean, called LeanSAT.
 
@@
 only reason about bitvector expressions where the width of the
-bitvectors involved is bounded. One of the topics I'd like to explore
-more is the connection between bitvectors and 2-adic numbers (which are
+bitvectors involved is bounded --- and even then, only if the bound is not too large.
+One of the topics I'd like to explore more is the connection between bitvectors and 2-adic numbers (which are
 infinite and thus coinductive in nature!) to build a decision procedure
@@
 
-Beyond being more satisfying, this arbitrary-width procedure is likely
-more performant than SAT solving, and thus relevant even if the end-user
-isn't necessarily interested in proving their property for all possible
-widths.
+Furthermore, during my earlier work on LeanMLIR, we've already encountered a 
+few other strategies for solving other classes of arbitrary-widt bitvector problems.
+So, even though deciding these problems in general is undecidable, we can identify
+specific classes of problems that do have decision procedures.
+
+I'm planning to write a paper about this work tentatively titled: "\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}", 
+in which we explore all of these different strategies, 
+classify exactly which class of expressions they solve, 
+and investigate how many real-world problems are already covered by these strategies.
+The aim is to submit this to PLDI, which has its call for papers around November.
 
@@
 verifying (low-level) MLIR dialects.
+
+This internship will thus provide me with exactly the corpus of real-world bitvector verification problems that I need to evaluate the strategies we'll discuss in the 
+aforementioned paper.
 
@@
 
+\subsubsection{Graph Regions \& Parametric
+Coinduction}\label{graph-regions-parametric-coinduction}
+
+\begin{quote}
+December '24 - June '25
+\end{quote}
+
+A \emph{Graph Region} in MLIR is, just like a regular region, a sequence
+of instructions with their results bound to variables. However, unlike
+normal regions, in a graph region the \emph{use} of a variable may
+precede it's definition. For example
+
+\begin{verbatim}
+^bb(%x : i32)
+  %y = comb.and %x %y : (i32, i32) -> (i32)
+\end{verbatim}
+
+Here, \texttt{y} is obtained as the bitwise and of the argument
+\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
+sense, but this style of definition is common in hardware
+representations, where this is interpreted as: ``the value of \texttt{y}
+in the \emph{current} clock cycle is computed by taking the bitwise and
+of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles.
+See https://circt.llvm.org/ for more details} To formally model this, we could denote variables as infinite streams of bitvectors,
+which gives us the value at every possible future clock cycle. Another
+example of a dialect with graph regions is the FSM, or Finite State
+Machine, dialect.
+
+Graph regions are thus inherently stateful, and modelling them is a
+particularly good use-case for coinductive types and coinductive
+reasoning. To facilitate easy verification of coinductive properties in
+this setting, I would like to investigate supporting parametrized
+coinductive proofs~\cite{hurPowerParameterizationCoinductive} in
+the coinductive types framework.
+
+Once we've got this this, I'd like to actually write a paper about the coinduction library, using this work on graph regions as a case study.
+The goal would be to submit this to POPL 25.
+
+\subsection{LeanMLIR}\label{leanmlir}
+
+Finally, I'd like to explore further uses of coinduction in the formalization of compiler IRs.
+
+\subsubsection{Non-termination}\label{non-termination}
+
+\begin{quote}
+June '25 - September  '25
+\end{quote}
+
+In the status report, we discussed how the framework was generalized to
+work in arbitrary monads. But, to actually define the semantics of an
+MLIR dialect which allows for infinite loops, we need a concrete monad
+that can support this behaviour.
+
+We could, of course, model non-termination using the \texttt{Option}
+monad, where divergence is simply modelled as \texttt{Option.none}. Such
+semantics are necessarily non-computable, losing the very nice property
+that the formalization of a dialect in LeanMLIR also gives a reference
+implementation for an interpreter.
+
+However, coinduction is actually a very good fit for this problem as well.
+Therefore, I'd like to use my library for coinductive types to define a
+coinductive monad which supports computable infinite behaviours, drawing
+inspiration from the interaction tree library for coq.~\cite{xiaInteractionTreesRepresenting2020}
+This addresses a
+weakness in LeanMLIR while simultaneously giving a nice case study for
+the usability of the coinduction library.
+
 \subsubsection{Runtime behaviour}\label{runtime-behaviour}
 \begin{quote}
-Jan '25 - May '25
+September '25 - May '25
 \end{quote}
+
+Once we start using this interaction tree datastructure in LeanMLIR, though, 
+we will likely run into performance problems if we try to actually execute our computable semantics.
 
@@
 of state machines (or, more generally, coalgebras).
-
-\subsection{LeanMLIR}\label{leanmlir}
-
-Finally, we return to the main objective: formalizing compiler IRs.
-
-\subsubsection{Non-termination}\label{non-termination}
-
-\begin{quote}
-Apr '25 - June '25
-\end{quote}
-
-In the status report, we discussed how the framework was generalized to
-work in arbitrary monads. But, to actually define the semantics of an
-MLIR dialect which allows for infinite loops, we need a concrete monad
-that can support this behaviour.
-
-We could, of course, model non-termination using the \texttt{Option}
-monad, where divergence is simply modelled as \texttt{Option.none}. Such
-semantics are necessarily non-computable, losing the very nice property
-that the formalization of a dialect in LeanMLIR also gives a reference
-implementation for an interpreter.
-
-Therefore, I'd like to use my library for coinductive types to define a
-coinductive monad which supports computable infinite behaviours, drawing
-inspiration from the interaction tree library for coq.
-\cite{xiaInteractionTreesRepresenting2020} This addresses a
-weakness in LeanMLIR while simultaneously giving a nice case study for
-the usability of the coinduction library.
-
-\subsubsection{Graph Regions \& Parametric
-Coinduction}\label{graph-regions-parametric-coinduction}
-
-\begin{quote}
-June '25 - Sep '25
-\end{quote}
-
-A \emph{Graph Region} in MLIR is, just like a regular region, a sequence
-of instructions with their results bound to variables. However, unlike
-normal regions, in a graph region the \emph{use} of a variable may
-precede it's definition. For example
-
-\begin{verbatim}
-^bb(%x : i32)
-  %y = comb.and %x %y : (i32, i32) -> (i32)
-\end{verbatim}
-
-Here, \texttt{y} is obtained as the bitwise and of the argument
-\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
-sense, but this style of definition is common in hardware
-representations, where this is interpreted as: ``the value of \texttt{y}
-in the \emph{current} clock cycle is computed by taking the bitwise and
-of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{
-  This example is simplified, in reality you'd have to specify a
-  register instruction to control when a value is stored across clock
-  cycles. See https://circt.llvm.org/ for more details}
-To formally model this, we could denote variables as infinite streams of bitvectors,
-which gives us the value at every possible future clock cycle. Another
-example of a dialect with graph regions is the FSM, or Finite State
-Machine, dialect.
-
-Graph regions are thus inherently stateful, and modelling them is a
-particularly good use-case for coinductive types and coinductive
-reasoning. To facilitate easy verification of coinductive properties in
-this setting, I would like to investigate supporting parametrized
-coinductive proofs~\cite{hurPowerParameterizationCoinductive} in
-the coinductive types framework.
 

### 2024-06-25 paper.tex

--- 2024-06-25 paper.tex
@@
 
-Now, formal verification has been done before. Most notable is the
-CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}. Yet, the vast majority of
+Now, formally verified compilation has been done before.
+Most notable is the CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}.
+Yet, the vast majority of
 code today is still being compiled with non-verified compilers. We ask,
@@
 
-\subsection{LeanMLIR}\label{leanmlir}
-
-Finally, we return to the main objective: formalizing compiler IRs.
-
-\subsubsection{Non-termination}\label{non-termination}
-
-\begin{quote}
-Apr '25 - June '25
-\end{quote}
-
-In the status report, we discussed how the framework was generalized to
-work in arbitrary monads. But, to actually define the semantics of an
-MLIR dialect which allows for infinite loops, we need a concrete monad
-that can support this behaviour.
-
-We could, of course, model non-termination using the \texttt{Option}
-monad, where divergence is simply modelled as \texttt{Option.none}. Such
-semantics are necessarily non-computable, losing the very nice property
-that the formalization of a dialect in LeanMLIR also gives a reference
-implementation for an interpreter.
-
-Therefore, I'd like to use my library for coinductive types to define a
-coinductive monad which supports computable infinite behaviours, drawing
-inspiration from the interaction tree library for coq.
-\cite{xiaInteractionTreesRepresenting2020} This addresses a
-weakness in LeanMLIR while simultaneously giving a nice case study for
-the usability of the coinduction library.
-
-\subsubsection{Graph Regions \& Parametric
-Coinduction}\label{graph-regions-parametric-coinduction}
-
-\begin{quote}
-June '25 - Sep '25
-\end{quote}
-
-A \emph{Graph Region} in MLIR is, just like a regular region, a sequence
-of instructions with their results bound to variables. However, unlike
-normal regions, in a graph region the \emph{use} of a variable may
-precede it's definition. For example
-
-\begin{verbatim}
-^bb(%x : i32)
-  %y = comb.and %x %y : (i32, i32) -> (i32)
-\end{verbatim}
-
-Here, \texttt{y} is obtained as the bitwise and of the argument
-\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
-sense, but this style of definition is common in hardware
-representations, where this is interpreted as: ``the value of \texttt{y}
-in the \emph{current} clock cycle is computed by taking the bitwise and
-of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{
-  This example is simplified, in reality you'd have to specify a
-  register instruction to control when a value is stored across clock
-  cycles. See https://circt.llvm.org/ for more details}
-To formally model this, we could denote variables as infinite streams of bitvectors,
-which gives us the value at every possible future clock cycle. Another
-example of a dialect with graph regions is the FSM, or Finite State
-Machine, dialect.
-
-Graph regions are thus inherently stateful, and modelling them is a
-particularly good use-case for coinductive types and coinductive
-reasoning. To facilitate easy verification of coinductive properties in
-this setting, I would like to investigate supporting parametrized
-coinductive proofs~\cite{hurPowerParameterizationCoinductive} in
-the coinductive types framework.
-
 % \subsection{Stretch Goal: Iris}\label{stretch-goal-iris}
-
+% 
 % \begin{quote}

### 2024-06-25 paper.tex
fixes

@@
 
-Lopes et al.~propose \emph{translation validation} as a way to scale
+Lopes\etal{} propose \emph{translation validation} as a way to scale
 formal verification~\cite{lopesAlive2BoundedTranslation2021}.
@@
 
-For example, Xia et al.~used interaction trees, a coinductive type, in
+For example, Xia\etal{} used interaction trees, a coinductive type, in
 their Coq formalization of recursive and impure programs~\cite{xiaInteractionTreesRepresenting2020}.
@@
 in the \emph{current} clock cycle is computed by taking the bitwise and
-of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles.
-See https://circt.llvm.org/ for more details} To formally model this, we could denote variables as infinite streams of bitvectors,
+of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles.
+See https://circt.llvm.org/ for more details}.
+To formally model this, we could denote variables as infinite streams of bitvectors,
 which gives us the value at every possible future clock cycle. Another
@@
 coinductive monad which supports computable infinite behaviours, drawing
-inspiration from the interaction tree library for coq.~\cite{xiaInteractionTreesRepresenting2020}
+inspiration from the interaction tree library for Coq~\cite{xiaInteractionTreesRepresenting2020}.
 This addresses a

### 2024-06-25 paper.tex
fix timeline

@@
 certain high-risk domains, where traditionally optimizations would be
-turned off entirely,~\cite{kastnerCompCertPracticalExperience} the
-vast majority of C code is thus still compiled with non-verified
+turned off entirely~\cite{kastnerCompCertPracticalExperience},
+the vast majority of C code is thus still compiled with non-verified
 compilers, which studies have shown to be prone to miscompilations~\cite{yangFindingUnderstandingBugs2011}.
@@
 particularly suited for this purpose because of it's community of
-mathematicians developing a comprehensive mathematical library,~\cite{themathlibcommunityLeanMathematicalLibrary2020} 
+mathematicians developing a comprehensive mathematical library~\cite{themathlibcommunityLeanMathematicalLibrary2020},
 and because of its highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.
@@
 \begin{quote}
-September '25 - May '25
+September '25 - Jan '26
 \end{quote}

### 2024-06-25 paper.tex
grammar/spelling pass

@@
 to compile, e.g., C, C++, Fortran, Rust, and Swift. Verifying the entire
-LLVM codebase, comprising of a few million lines of C++ code, is
+LLVM codebase, consisting of a few million lines of C++ code, is
 infeasible. Instead, we follow the same approach as Lopes\etal{} by
@@
 because it's an end-to-end verified monolith, CompCert is not
-competitive compared to traditional non-verified compilers. Outside of
+competitive compared to traditional non-verified compilers. Outside
 certain high-risk domains, where traditionally optimizations would be
@@
 
-Alive is able to leverate SMT solvers because the semantics of LLVM IR
+Alive is able to leverage SMT solvers because the semantics of LLVM IR
 are given in terms of bitvectors, and SMT solvers understand the
@@
 much more expressible than SMT solvers. Among theorem provers, Lean is
-particularly suited for this purpose because of it's community of
+particularly suited for this purpose because of its community of
 mathematicians developing a comprehensive mathematical library~\cite{themathlibcommunityLeanMathematicalLibrary2020},
@@
 centred on finite state machines. Such automata are understood to be
-coinductive in nature, so mechanizing the semanantics of this dialect in
+coinductive in nature, so mechanizing the semantics of this dialect in
 Lean will likely require coinduction.
@@
 which is based on earlier
-work by Avigad et al that models coinductive as quotients of polynomial
+work by Avigad\etal{} that models coinductive as quotients of polynomial
 functors (QPFs)~\cite{avigadDataTypesQuotients2019}.
@@
 the kernel to understand coinduction and ensure the kernel recognizes
-the desired definitonal equalities. However, modifying the trusted
+the desired definitional equalities. However, modifying the trusted
 kernel carries a large burden of proof, since the changes could
@@
 property\footnote{The subject reduction property states that reducing a program does not change its type}~\cite{sozeauCorrectCompleteType}.
-Our coinductives-as-a-library approach requires no new axioms nor
+Our coinduction-as-a-library approach requires no new axioms nor
 changes to the kernel, and is thus guaranteed not to change any meta
@@
 
-However, we've focused on dialects with pure (i.e., side-effect free)
+However, we've focused on dialects with pure (i.e., side effect free)
 semantics. Even though we \emph{are} able to verify a lot of interesting
 rewrites in high-level MLIR dialects, we do need to be able to reason
-about side-effect if we want to capture the low-level dialects as well,
+about side effect if we want to capture the low-level dialects as well,
 as a reviewer of the paper rightly pointed out.
 
-To this end, we've generalised the framework so that a dialect's
+To this end, we've generalized the framework so that a dialect's
 semantics may be given in a user-defined monad, and proven that applying
 a semantics-preserving pure rewrite (i.e., where the operations being
-rewritten are all side-effect free) inside of a side-effecting program
+rewritten are all side effect free) inside a side-effecting program
 preserves semantics of the program as a whole. However, this
-generalisation still has a lot of rough edges, and further work is
+generalization still has a lot of rough edges, and further work is
 needed before we can claim our framework is versatile enough to deal
-with side-effects cleanly.
+with side effects cleanly.
 
@@
 
-As mentioned, a big reason for us to use Lean is it's community. As
+As mentioned, a big reason for us to use Lean is its community. As
 such, I've been heavily involved in the redesign of Lean's API for
@@
 
-This automation is crucial, because with it we can provide push-button
-automation for non-expert users on the same level as Alive, for dialects
-whose semantics use bitvectors, while also still having the full
-expressivity needed for dialects that use more complex mathematical
+This automation is crucial, 
+because it brings the user experience for non-experts to the same level as Alive, 
+for dialects whose semantics use bitvectors, 
+while still having the full expressivity needed for dialects that use more complex mathematical
 objects. Thus, addressing the main drawback of using a theorem prover.
@@
 Furthermore, during my earlier work on LeanMLIR, we've already encountered a 
-few other strategies for solving other classes of arbitrary-widt bitvector problems.
+few other strategies for solving other classes of arbitrary-width bitvector problems.
 So, even though deciding these problems in general is undecidable, we can identify
@@
 Furthermore, I plan to continue this work on bitvector automation in
-Lean during my internship at Amazon \footnote{ Assuming I'll get
+Lean during my internship at Amazon\footnote{Assuming I'll get
   permission to work away, of course}. They are particularly interested
@@
 principles, which is then compiled into an application of this
-corecursion princple. This will involve some software engineering,
+corecursion principle. This will involve some software engineering,
 dealing with Lean's extensible syntax, plus a good amount of research,
@@
 productive is paramount to our logical system remaining consistent. In
-fact, because of our coinductives-as-a-library approach we are unable to
+fact, because of our coinduction-as-a-library approach we are unable to
 define corecursive functions without proving they are productive, and
@@
 \emph{guarded} productivity checker. That is, this first implementation
-will only allow only functions where each corecursive call occurs behind
+will only allow functions where each corecursive call occurs behind
 a constructor. This check is known to incomplete, and more complicated
@@
 now --- the initial implementation of corecursion in Coq also allowed
-only guarded corecursion \cite{gimenezTutorialRecursiveTypes1998}.
+only guarded corecursion~\cite{gimenezTutorialRecursiveTypes1998}.
 Of course, if there is still time left in the internship after this goal
@@
 normal regions, in a graph region the \emph{use} of a variable may
-precede it's definition. For example
+precede its definition. For example
 
@@
 in the \emph{current} clock cycle is computed by taking the bitwise and
-of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles.
+of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles.
 See https://circt.llvm.org/ for more details}.
@@
 
-Once we've got this this, I'd like to actually write a paper about the coinduction library, using this work on graph regions as a case study.
+Once we've got this, I'd like to actually write a paper about the coinduction library, using this work on graph regions as a case study.
 The goal would be to submit this to POPL 25.
@@
 Finite trees are easily encoded in Lean, but representing a finite tree
-is a bit more tricky. The current construction works by representing an
+is a bit more tricky. The current construction works by representing a
 non-wellfounded tree as a function from a natural number \(n\) to the

### 2024-06-25 paper.tex
fix quotes

@@
 
-I'm planning to write a paper about this work tentatively titled: "\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}", 
+I'm planning to write a paper about this work tentatively titled: ``\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}'', 
 in which we explore all of these different strategies, 

### 2024-06-25 paper.tex
minor fixes

@@
 to know which functions are \emph{productive} in our formalization, and
-to automatically recognize (a well-defined sub-class of) these
+to automatically recognize (a well-defined subclass of) these
 productive corecursive functions.

### 2024-12-05 paper.tex
Update paper.tex
document my internship experience

@@
 
-I'm planning to write a paper about this work tentatively titled: ``\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}'', 
-in which we explore all of these different strategies, 
-classify exactly which class of expressions they solve, 
-and investigate how many real-world problems are already covered by these strategies.
-The aim is to submit this to PLDI, which has its call for papers around November.
+This culminated in a paper titled: ``\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}'', 
+which we submitted to PLDI.
 
@@
 
-Furthermore, I plan to continue this work on bitvector automation in
-Lean during my internship at Amazon\footnote{Assuming I'll get
-  permission to work away, of course}. They are particularly interested
-in automatically reasoning about ARM assembly code, which closely aligns
-with our goal of automating bitvector reasoning to then use for
-verifying (low-level) MLIR dialects.
+At Amazon, I wrote a scalable symbolic simulator for straight-line, memory-free Arm machine code, in the context of LNSym, 
+a toolkit for the verification of cryptographic hash functions and block ciphers written in Arm assembly.
+Before my internship, a prototype symbolic simulator existed, but it was limited in features and did not scale well.
+My reimplemented version tracks more information of the simulated execution, 
+and is capable of simulating more than 400 instructions of the SHA512 litmus test we benchmarked against before hitting Lean's default time-out limits.
+This is about 10x more instructions than the prototype was able to simulate given the same time-out limits.
 
-This internship will thus provide me with exactly the corpus of real-world bitvector verification problems that I need to evaluate the strategies we'll discuss in the 
-aforementioned paper.
+The prototype relied heavily on Lean's builtin simplifier. 
+I replaced this with bespoke metacode, that was able to exploit the specific shape of the problem we were solving.
+Firstly, I wrote preprocessing that allowed us to treat the operational semantic model, where operations are given semantics as functions that write to state, 
+in an axiomatic-semantics way, by having theorems that characterizes operations in terms of reads from the resulting state.
+Secondly, I wrote a bespoke data-structure and used it selectively reflects parts of the proof goal.
+We explicitly decoupled this meta-code needed to scale symbolic simulation from the definition of the semantic model,
+so that the model remains extensible even by people who are not experts in formal verification, which was an explicit design goal.
+
+This project, like some of my other research, was also an experiment in the usability of Lean as a platform to verify software.
+In particular, we generated very large proofs, and exposed, amongst others, a fundamental scalability problem in Lean:
+The time Lean needs to assemble a proof after a tactic executes scales roughly by the length of the proof times the number of binders in the proof.
+In the larger benchmarks, LNSym would generate proofs involving tens of thousands of binders,
+resulting in the majority of time needed to check the proof was actually spent in this proof-assembling step.
+We reported this to the Lean core team, who are actively working on a solution.
+Furthermore, the semantic model of Arm machine code used relies heavily on bitvectors, thus, we also used the proof obligations resulting from the improved symbolic simulator to stress-test LeanSAT.
+I've reported these problems, and submitted patches for some, contributing to the greater vision of improving the usability of Lean.
+
+Besides the concrete contribution of the code I wrote, I also documented my experience,
+provifing an explanation of the various optimizations I tried, and which worked or didn't worked,
+resulting in a clear roadmap for the LNSym team to iterate on.
 

### 2025-01-31 notes/draftv2.tex
move new phd plan to .tex file

\section{Scalable and Modular Verified Compilation via Automated Proofs
for Declarative
Transformations}\label{scalable-and-modular-verified-compilation-via-automated-proofs-for-declarative-transformations}

A fully end-to-end verified \emph{mainstream} compiler with a small
trusted code base remains somewhat of an unattainable holy grail.
End-to-end verified compilers do exist (e.g., CompCert
{[}{[}@leroyCompCertFormallyVerified{]}{]} and CakeML
{[}{[}@kumarCakeMLVerifiedImplementation2014{]}{]}), but the extra
verification burden makes compiler development challenging, relegating
such compilers to specific high-assurance niches. On the other hand,
there are tools which can fully automatically prove the correctness of a
specific compiler transformation (e.g., Alive
{[}{[}@lopesAlive2BoundedTranslation2021{]}{]}, which has gained
widespread adoption in the LLVM community, and
{[}{[}@armstrongIslaIntegratingFullScale2021{]}{]}), but such tools
generally rely on SMT solvers to achieve their push-button automation,
making them unsuited in our quest for a minimal trusted code base. I
wish to investigate a middle road: can we verify the core transformation
algorithm separately from specific transformation instances, and then
develop proof automation for fully automatically showing local
correctness of a transformation in a way that composes into a proof of
global correctness?

Specifically, I'll develop two core components of a verified compiler:
an optimisation pass based on peephole optimisations and a code
generator that lowers the compiler IR (Intermediate Representation)
produced by the optimiser into assembly. Of course, a fully end-to-end
compiler would need many more components, such as a verified front-end
that parses source code and transforms it into semantically equivalent
compiler IR, however such components will be out-of-scope for my PhD.

To ensure a minimal trusted code base we'll develop our proofs and proof
automation in Lean. Besides a modern theorem prover, Lean is also aiming
to be a general purpose functional programming language. It is certainly
not the first theorem prover that allows its users to extract executable
code from a formalization, but Lean goes several steps further: the Lean
compiler is itself in large parts written in Lean. Furthermore, proof
automation for Lean is also written in Lean itself. On the one hand,
Lean's rich logic enable us to finely model the semantics of programs.
On the other, Lean itself aims to be a language that is suited for
large-scale software, which necessitates a modern optimising compiler
backend, but, Lean also has a feature that enables the use of
\emph{compiled} decision procedures in proofs. Using this feature makes
the optimiser part of the trusted codebase for our proofs, and it is
thus crucial for the logical soundness of our system that
mis-compilations do not occur. Combined, this makes Lean a compelling
platform for verified compilation.

For the optimisation pass, we choose to operate on LLVM IR --- the same
compiler IR used by Clang, Swift and Rust. Although others have
formalised the semantics of LLVM IR in a proof assistant before
({[}{[}@zhaoFormalVerificationSSABased{]}{]},
{[}{[}@zakowskiModularCompositionalExecutable2021{]}{]}), those efforts
included no proof automation. This will to our knowledge be the first
effort to provide fully automated proofs for correctness of LLVM
optimisations in a way that doesn't have to trust an external solver (as
Alive does {[}{[}@lopesAlive2BoundedTranslation2021{]}{]}). Removing
this dependency is important as SMT solvers are large, complex pieces of
software with a history of bugs
({[}{[}@brummayerFuzzingDeltadebuggingSMT2009{]}{]},
{[}{[}@mansurDetectingCriticalBugs2020{]}{]}).

The optimisation pass will be instantiated with a set of \emph{peephole
optimisations}, consisting of program patterns to match for and
corresponding program patterns to rewrite to. Lean-MLIR
({[}{[}@bhatVerifyingPeepholeRewriting2024{]}{]}), a project in
development at my research group which I've contributed to, currently
has a prototype of such an optimiser. However, the current version is
only able to apply \emph{pure} optimisations. I wish to expand this to
an optimiser that can apply optimisations involving side-effect such as
memory loads/stores, undefined behaviour, or potentially non-terminating
control flow, and, crucially, provide push-button proof automation for
such side-effecting optimisations, using symbolic execution.

After optimisation, the next phase will be to generate assembly from our
optimised IR and, crucially, to verify that the generated assembly has
the same semantics as the optimised IR. For maximum trust, we'd like to
use authoritative models of the ISA semantics of whichever platform we
are compiling for to give semantics to the assembly program. Sail
({[}{[}@armstrongISASemanticsARMv8a2019{]}{]}) provides such models: the
Sail model for Arm-A is automatically derived from the official ASL
reference, and their RISC-V model has been adopted by the RISC-V
Foundation. We'll phrase the code generator in a similar way: by having
fragments of compiler IR to match for, and corresponding fragments of
assembly to rewrite them with. Still, code generation presents some
unique challenges. Firstly, although the transformations themselves
should be simpler, the Sail semantic model will be much larger than our
handwritten LLVM semantics. Secondly, we have to deal with semantics of
different languages (the compiler IR and the assembly language). It is
likely that proof automation for code generation will thus need a
slightly different design than proof automation for peephole
optimisations.

Finally, to model the semantics of LLVM IR in a modular way, we'd like
to follow Zakowski et
al.~({[}{[}@zakowskiModularCompositionalExecutable2021{]}{]}) and use
interaction trees ({[}{[}@xiaInteractionTreesRepresenting2020{]}{]}).
Unfortunately, interaction trees are \emph{coinductive} in nature, to
capture possibly non-terminating programs, and Lean currently lacks
support for coinductive type. Therefore, I'd like to continue the work
that I started in own master's thesis
{[}{[}@keizerImplementingDefinitionalCo{]}{]}, which is based on earlier
work by Avigad et al.~({[}{[}@avigadDataTypesQuotients2019{]}{]}), and
develop a Lean library for coinductive types.

\textbf{My core contributions are as follows}: * A proof-producing
translation validation tool for compiler optimizations, which works on a
higher-level IR with relatively straightforward formalized semantics,
but is able to analyse complex transformations, including side effects.
* A proof-producing translation validation tool for code generation
using large-scale authoritative ISA models. Here the transformations are
relatively simple, but the complexity of the instruction semantics
presents unique challenges. * Finally, we develop technical
infrastructure needed for both of these tools in the form of coinductive
types support in Lean.

Note that the benefits of a verified compilation pipeline go far beyond
simply reducing the chance of bugs. We believe it will lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimisation to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.

Lean, again, presents a very interesting opportunity because of it's
\emph{extensibilty}: Lean already has an extensible syntax and a
powerful meta-programming API. In fact, a lot of features that are
considered part of Lean's core are actually implemented using the same
meta-programming facilities that are available to libraries (we'll use
this extensively in ch.~3). However, we imagine a future where this
extensibility could extend to the backend of the compiler also, such
that a library writer might implement an algorithm in a functional style
that is easy to reason about, but also provide an optimized assembly
\emph{and} a proof that the assembly is equivalent. Traditionally,
hard-coded assembly in a library has been cause for increased scrutiny.
Optimizing is hard, do we really trust the author of this library to
have done it correctly? Verification means we no longer have to trust,
instead we get to have fearless optimization!

\subsection{HIgh-Assurance Translation Validation for Compiler
Optimizations}\label{high-assurance-translation-validation-for-compiler-optimizations}

LLVM IR is an intermediate representation that sits just above the level
of abstraction of assembly, and has proven a successful platform to
perform optimisations on before finally lowering to platform-specific
assembly. LLVM has been used as a back-end for many mainstream
compilers, such as Clang, Rust and Swift.

A crucial property of LLVM IR, is that it is in static single assignment
(SSA) form: rather than operating on a fixed set of registers, LLVM IR
has variables that are assigned exactly once. This form makes it easier
to apply peephole rewrites, as we can simply match for a pattern along
what is called the def-use chain, even if there are unrelated
instructions in between.

Now, although LLVM is a bit higher level than assembly, it's still
low-level and, e.g., features unstructured control flow. The LLVM
community realised that having more structure available would enable
even more optimisations to be done purely locally (i.e., as peephole
rewrites). Thus, MLIR (Multi Level Intermediate Representation) was
born: a framework to define compiler IRs at various levels of
abstraction. MLIR mandates that IRs are in SSA form, and takes care of
syntactic overhead. We can instantiate the dialect with a specific set
of operations, called a dialect. Generally, a copmiler IR is made up of
a combination of such dialects. For this chapter, we shall focus on an
IR that contains bitvector arithmetic (a fragment of the \texttt{arith}
dialect) memory manipulations (\texttt{ptr} dialect) and structured
control flow (i.e., for and while loops, \texttt{scf} dialect). This is
still fairly low level, but the use of structured control flow puts us
just above the level of abstraction of LLVM IR.

\subsubsection{Lean-MLIR Today}\label{lean-mlir-today}

Lean-mlir is a framework for modelling the \emph{semantics} of MLIR
dialect in Lean, developed by my research group
{[}{[}@bhatVerifyingPeepholeRewriting2024{]}{]}, focusing on the
\emph{pure} (i.e., side-effect-free) fragment of MLIR dialects.

In particular, lean-mlir currently has a model of just the bitvector
arithmetic operations of LLVM IR, and a verified peephole rewriter that
can apply a pure peephole rewrite (i.e., neither the program fragment
being matched on nor the fragment being rewritten to contain
side-effecting operations) and, assuming the local correctness of every
rewrite, we've shown the rewriter preserves the global semantics.

Furthermore, the Lean FRO has, in collaboration with my research group
and myself, been developing proof automation for deciding the equality
of fixed-width bitvector expressions {[}CITATION NEEDED{]}. In
particular, this proof automation allows us to fully automatically
decide the local correctness of \emph{pure} optimisations. Note that
although this tactic employs an external (certificate-producing) SAT
solver, it also includes a verified certificate checker, so that the
final proofs do \emph{not} rely on the correctness of the solver. That
said, the checker would be too slow if ran inside the Lean kernel, so to
get decent performance, this tactic does rely on the previously
mentioned feature to run the compiled version of a Lean, thus expanding
the trusted code base to include the Lean compiler.

\subsubsection{Proof Automation for Side
Effects}\label{proof-automation-for-side-effects}

I wish to expand upon this first by modelling all operations of the
dialect mentioned before. The bitvector arithmetic fragment of the
previous section is basically the same as \texttt{arith}, and the
framework also already models part of \texttt{scf} (in particular, it
models counting for-loops, but not while-loops with arbitrary conditions
yet), so the main missing piece is a model of memory and memory
operations. To make this easier, I'll focus purely on single-threaded
code, foregoing the need for complicated memory models.

The main innovation in this project will be to build proof automation in
Lean for translation validation on this compiler IR. That is, given a
source and target program (potentially including holes), will
automatically decide the correctness of the transformation. To do this
in the presence of side-effects, I'll build a symbolic simulator.

\subsubsection{Refinement and Undefined
Behaviour}\label{refinement-and-undefined-behaviour}

So far, we've talked about semantic equivalence as the condition for
correctness of an optimisation, but in reality it's a bit more
complicated because of the notion of undefined behaviour (UB). In short,
if a program has UB, it's legal to change the program to anything.
Conversely, if a source program is fully defined, then it's illegal to
introduce UB in the target. The corresponding relation between programs
is called \emph{refinement}. \#\#\# Related Work

We can broadly categorise earlier efforts to apply formal methods to
LLVM into those that formalise the semantics of LLVM, but perform manual
proofs (e.g., VeLLVM {[}{[}@zhaoFormalVerificationSSABased{]}{]},
{[}{[}@zhaoFormalizingLLVMIntermediate2012{]}{]} or work by Zakowski et
al.~{[}{[}@zakowskiModularCompositionalExecutable2021{]}{]}) and tools
like Alive ({[}{[}@lopesAlive2BoundedTranslation2021{]}{]}) that perform
fully automated translation validation, but rely on SMT solvers.
AliveInLean is a prototype that shows how to increase trust without
sacrificing automation, but this project still trusts the SMT solver,
and does not support reasoning about memory
{[}{[}@leeAliveInLeanVerifiedLLVM2019{]}{]}.

More broadly, CompCert ({[}{[}@leroyCompCertFormallyVerified{]}{]}) and
CakeML ({[}{[}@kumarCakeMLVerifiedImplementation2014{]}{]}) are examples
of fully end-to-end verified compilers. More recent work by Mullen et
al.~{[}{[}@mullenVerifiedPeepholeOptimizations2016{]}{]} argues for
peephole rewriting as a way to lower proof burden in CompCert
development, but they still prove these rewrites manually.

Katamaran {[}{[}@keuchelVerifiedSymbolicExecution2022{]}{]} is a
symbolic simulation tool written in a different ITP. However, this
project focusses on being able to run the simulator outside of the ITP
environment, hence eschewing meta-programming. Our main purpose is
modular proof automation, so being tied to Lean is no problem for us,
allowing us to make different implementation choices.

\subsection{Scaling High-Assurance Translation Validation to
Authoritative ISA
Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models}

After performing optimisations on the level of our compiler IR, we need
to lower this representation to assembly. We'll do this in a very
similar way, by phrasing instruction selection as rewrites that match
for a compiler IR program fragment and rewrite it to an assembly program
fragment. By following a similar approach, we can hopefully reuse some
of the infrastructure we developed in the previous chapter.

\subsubsection{The RISC-V Dialect}\label{the-risc-v-dialect}

Lopoukhine et al {[}CITATION NEEDED{]} have developed a set of MLIR
dialects that together model RISC-V assembly. Representing our target
for code generation as an MLIR dialect allows us to use Lean-MLIR,
enabling reuse of the core datastructures provided by the framework, but
the benefits extend much further! Namely, depending on which of the
dialects provided by Lopoukhine et al.~we use, our representation of
assembly can (a) be either register allocated, or as an assembly-like
SSA IR, and (b) use either structured (for/while loops) or unstructured
control flow (branches).

We'll use these progressively lower level IRs as targets for our code
generator, explicitly separating the instruction selection pass, which
we'll implement via a simple rewriter, from a register allocation pass.
Finally, recall that we chose to stick to structured control flow in the
optimisation pass to preserve structure. We'll do the same in the code
generator, postponing the lowering from structured into unstructured
control flow to the last moment. \#\#\# Semantics for RISC-V

So far, we've eliminated any reliance on SMT solvers in pursuit of a
smaller trusted code base. But recall that the semantics model of
whichever language we choose to prove things about are also a crucial
part of the trust story. In the end, if we run our compiler on actual
hardware, then a proof of correctness is only good so long as the actual
hardware behaves the same as our model of the instruction set
architecture (ISA).

One could attempt to formally verify the hardware as well, but that is
well beyond the scope of my PhD. Instead, I'd like to use Sail models as
an authoritative source of truth for ISA semantics
{[}{[}@armstrongISASemanticsARMv8a2019{]}{]}. Sail is a domain specific
language for modelling ISA semantics. Various models exists, but in
particular, the Arm-v8 Sail model has been automatically derived from
the Arm-internal specification
{[}{[}@reidTrustworthySpecificationsARM{]}{]} and the RISC-V model has
been adopted by RISC-V International
{[}{[}@sammlerIslarisVerificationMachine2022{]}{]}. Using these models
as the source of our semantics gives us trust that any discrepancy
between our formalization and actual hardware likely is a bug in the
hardware!

The Sail project currently provides a translation from Sail to Coq,
which has been used previously to formally reason about ISAs. Luckily, a
translation from Sail to Lean is currently being developed, following a
similar strategy to add a backend to sail that emits Lean code---as
opposed to embedding a model of Sail in Lean.

Of course, the Sail semantics will operate on traditional assembly, with
registers allocated and unstructured branches. This should not be a
problem, as lowering a single instruction at any of the progressive
levels of abstraction down to regular assembly should not be difficult.

\subsubsection{Translation Validation}\label{translation-validation}

Again, the core contribution will be proof automation. By treating
instruction selection as a sequence of rewrites, we merely need to
verify that each of these rewrites individually preserves semantics.

This seems like much the same problem that we tackled in the previous
chapters, but there are some subtleties. For one, the rewrites
themselves should be simpler, with many rewrites having only a single
instruction as source and target. However, a Sail-derived semantics
model will be much larger than the hand-written model we'll be using for
the compiler IR, necessitating many more reasoning steps to simplify the
semantics for a single instruction. Many of the details that make the
Sail model so large might be irrelevant in our context, so I'll
investigate ways to simplify the model under specific assumptions. For
example, by assuming we're only compiling user-level programs, we might
be able to simplify away any complexity to do with kernel-level
instructions in the model.

\paragraph{LNSym}\label{lnsym}

During my internship at Amazon, I developed a symbolic simulator for a
simplified fragment of Arm assembly programs, using a hand-written model
of the ARM ISA semantics. My internship focused primarily on making this
symbolic simulator scale to being able to simulate hundreds of
instructions within a reasonable timeframe. I'll be using this symbolic
simulator as a basis for the translation validation tool I aim to
develop, adapting it to the slightly different design constraints in
this context.

\subsubsection{Related Work}\label{related-work}

\begin{quote}
{[}!TODO{]} Related work: Isla/Islaris? Probably introduce Sail
properly?
\end{quote}

\subsection{A Codatatypes Library for
Lean}\label{a-codatatypes-library-for-lean}

In previous chapters, we've glossed over how, exactly, I'll give
semantics to the side-effectful version of our compiler IR. Zakowski et
al.~have written about the the benefits of an interaction tree (ITree)
based semantics {[}{[}@zakowskiModularCompositionalExecutable2021{]}{]},
{[}{[}@xiaInteractionTreesRepresenting2020{]}{]}, so I'd like to use
ITrees also. Sadly, Lean does not have support coinductive types. I'd
like to address this, and build an ergonomic Lean library for ITrees.

\subsubsection{Quotients of Polynomial
Functors}\label{quotients-of-polynomial-functors}

Luckily, we can encode coinductives in Lean's existing logic, using
so-called \emph{quotients of polynomial functors} (QPFs). Avigad et
al.~have written about the low-level construction
{[}{[}@avigadDataTypesQuotients2019{]}{]}, upon which I expanded in my
Master's thesis by building a higher-level framework for defining
coinductive types, called QPFTypes. This framework is far from perfect,
and I'll spend some time fixing bugs in the framework. Still, the main
goal will be simply to define interaction trees; not to perfectly polish
the framework.

\subsubsection{Corecursive Functions}\label{corecursive-functions}

The QPFTypes framework currently only has facilities for defining types;
to define corecursive functions that operate over ITrees we only have
the low-level corecursion principle to work with. However, in their
original development of ITrees, Xia et al in fact provide a set of three
combinators which they claim make working with ITrees more tractable,
and remove the need for users to work directly with Coq's implementation
of corecursive function (which they claim is not compositional). As it
turns out, these combinators are rather similar to the corecursion
principle we have available, and the generic abstractions that William
developed on top of the corecursion principle. So, for our ITree library
in Lean, we don't have to worry about general corecursive functions; it
suffices to implement the three basic combinators that Xia et
al.~provide.

\subsubsection{Related Work}\label{related-work-1}

The QPF construction encodes coinductive types as a \emph{library},
instead of having to modify the logical system of Lean to support
coinduction natively --- an approach that is greatly aided by Lean's
excellent meta-programming capabilities. When looking at coinductive
types in other theorem provers, we distinguish between languages like
Isabelle {[}{[}@traytelCategoryTheoryBased{]}{]}, which follow the same
coinduction-as-a-library approach, and languages like Coq
{[}{[}@gimenezTutorialRecursiveTypes1998{]}{]}{[}{[}@gimenezApplicationCoinductiveTypes1996{]}{]}
or Agda which have modified their trusted kernels to support
coinduction.

Isabelle, in particular, is relevant because it's construction of
coinductive types, in terms of \emph{bounded natural functors} is
closely related to our construction in terms of QPFs.
{[}{[}@furerQuotientsBoundedNatural2022{]}{]} There are still new
challenges, however, because Lean's dependent types mean we have to
worry about the distinction between definitional equality (which is a
decidable, but weaker notion of equality) and propositional equality
(which captures more things that are intuitively equal, but requires
proof). Isabelle, in contrast, uses a weaker logical system where all
equalities are decidable.

In contrast, the coinduction-in-the-kernel approach would modify the
kernel to understand coinduction and ensure the kernel recognizes the
desired definitional equalities. However, modifying the trusted kernel
carries a large burden of proof, since the changes could compromise the
logical soundness of the entire system. In fact, the original
implementation of coinduction, called \emph{positive} coinduction, is
nowadays discouraged in favour of an alternative implementation, since
positive coinduction breaks the subject reduction property
\footnote{ The subject reduction property states that reducing a program
  does not change its type}{[}{[}@sozeauCorrectCompleteType{]}{]}. The
coinductives-as-a-library approach requires no new axioms nor changes to
the kernel, and is thus guaranteed not to change any meta properties of
Lean.

\subsection{Timeplan}\label{timeplan}

My PhD clock started in Michealmas term 2023, meaning my submission
window is between October 2026 and September 2027, or between 21 to 33
months from now. Picking the lower end of that timeframe gives exactly 7
months per chapter, leaving the final year for writing up the thesis.
\#\#\# High-Assurance Translation Validation for Compiler Optimizations
Targeted publication venue: PLDI

\begin{itemize}
\tightlist
\item
  Translation Validation for arithmetic \& bounded, structured control
  flow (2 months)
\item
  Translation Validation for memory side-effects (2 months)
\item
  Stress-test and improve scalability of proof automation (2 months)
\item
  Paper writing (1 month)
\end{itemize}

\subsubsection{A Codatatypes Library for
Lean}\label{a-codatatypes-library-for-lean-1}

Targeted publication venue: ITP

\begin{itemize}
\tightlist
\item
  Bugfix framework to accept ITree definition (2 months)
\item
  Define basic ITree API, including combinators (2 months)
\item
  Evaluate ITrees by expanding compiler IR semantics to infinite loops
  (2 months)
\item
  Paper writing (1 month)
\end{itemize}

\subsubsection{Scaling High-Assurance Translation Validation to
Authoritative ISA
Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models-1}

Targeted publication venue: PLDI

\begin{itemize}
\tightlist
\item
  Define lowering from compiler IR to assembly (2 months)
\item
  Translation validation for unstructured control flow (2 months)
\item
  Evaluate and improve scalability of SailToLean-generated ISA model (2
  months)
\item
  Paper writing (1 month)
\end{itemize}

### 2025-01-31 paper.tex
reword contributions as research questions
add criteria for success and risks and mitigations
add explicit back-references to relevant research question in each chapter

--- 2024-06-25 paper.tex
@@
                                       %% header instead of Full Title.
-\title{Towards a framework for reasoning about stateful compiler IRs 
-using coinductive reasoning.}                                      
+\title{Scalable and Modular Verified Compilation via Automated Proofs
+for Declarative
+Transformations.}                                      
 \subtitle{First Year Report}
@@
 
-\section{Introduction}
+A fully end-to-end verified \emph{mainstream} compiler with a small
+trusted code base remains somewhat of an unattainable holy grail.
+End-to-end verified compilers do exist (e.g., CompCert~\cite{leroyCompCertFormallyVerified} and CakeML~\cite{kumarCakeMLVerifiedImplementation2014}), but the extra
+verification burden makes compiler development challenging, relegating
+such compilers to specific high-assurance niches. On the other hand,
+there are tools which can fully automatically prove the correctness of a
+specific compiler transformation (e.g., Alive~\cite{lopesAlive2BoundedTranslation2021}, which has gained
+widespread adoption in the LLVM community), but such tools
+often rely on SMT solvers to achieve their push-button automation,
+making them unsuited in our quest for a minimal trusted code base. I
+wish to investigate a middle road: can we verify the core transformation
+algorithm separately from specific transformation instances, and then
+develop proof automation for fully automatically showing local
+correctness of a transformation in a way that composes into a proof of
+global correctness?
 
-Formal verification is crucial in ensuring the reliability and
-correctness of software systems. 
-Reasoning about compiler optimizations, in particular, is notoriously difficult, 
-and even minor bugs in widely-used compilers can lead to subtle, hard-to-debug miscompilations
-with far-reaching consequences. 
-This challenge makes the formal verification of compiler intermediate representations (IRs) a
-particularly compelling topic, 
-as it promises to enhance the dependability of the entire software development lifecycle.
+Specifically, I'll develop two core components of a verified compiler:
+an optimization pass based on peephole optimizations and a code
+generator that lowers the compiler IR (Intermediate Representation)
+produced by the optimizer into assembly. Of course, a fully end-to-end
+compiler would need many more components, such as a verified front-end
+that parses source code and transforms it into semantically equivalent
+compiler IR, however such components will be out-of-scope for my PhD.
 
-Now, formally verified compilation has been done before.
-Most notable is the CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}.
-Yet, the vast majority of
-code today is still being compiled with non-verified compilers. We ask,
-thus, how do we formally verify not just \emph{a} compiler, but, how do we
-scale formal verification to \emph{mainstream} compilers.
+To ensure a minimal trusted code base we'll develop our proofs and proof
+automation in Lean. Besides a modern theorem prover, Lean is also aiming
+to be a general purpose functional programming language. It is certainly
+not the first theorem prover that allows its users to extract executable
+code from a formalization, but Lean goes several steps further: the Lean
+compiler is itself in large parts written in Lean. Furthermore, proof
+automation for Lean is also written in Lean itself. On the one hand,
+Lean's rich logic enable us to finely model the semantics of programs.
+On the other, Lean itself aims to be a language that is suited for
+large-scale software, which necessitates a modern optimizing compiler
+backend, but, Lean also has a feature that enables the use of
+\emph{compiled} decision procedures in proofs. Using this feature makes
+the optimizer part of the trusted codebase for our proofs, and it is
+thus crucial for the logical soundness of our system that
+mis-compilations do not occur. Combined, this makes Lean a compelling
+platform for verified compilation.
 
-Take, for example, LLVM, a widely-used open-source compiler that is used
-to compile, e.g., C, C++, Fortran, Rust, and Swift. Verifying the entire
-LLVM codebase, consisting of a few million lines of C++ code, is
-infeasible. Instead, we follow the same approach as Lopes\etal{} by
-focussing on verifying that individual rewrites on a compiler IR preserve
-semantics~\cite{lopesAlive2BoundedTranslation2021}.
+For the optimization pass, we choose to operate on a compiler IR with the same semantics as LLVM.
+Although others have formalized the semantics of LLVM IR in a proof assistant before
+\cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
+included no proof automation. This will to our knowledge be the first
+effort to provide fully automated proofs for correctness of LLVM
+optimizations in a way that doesn't have to trust an external solver (as
+Alive~\cite{lopesAlive2BoundedTranslation2021} and even AliveInLean~\cite{leeAliveInLeanVerifiedLLVM2019} do). Removing
+this dependency is important as SMT solvers are large, complex pieces of
+software with a history of bugs~\cite{brummayerFuzzingDeltadebuggingSMT2009, mansurDetectingCriticalBugs2020}.
 
-Although LLVM is indeed used by the standard compilers for Rust and
-Swift, both also use other, language-specific, IRs to be able to do
-analyses and optimizations at a higher level of abstraction. In
-response, the MLIR (Multi-Level Intermediate Representation\footnote{Multi-Level
-  Intermediate Representation is the \emph{official} expansion of MLIR,
-  but other expansions are used by the community as well}) project has
-been created as a framework for creating domain-specific IRs (generally
-called a \emph{dialect}) at various levels of abstraction.
+% The next paragraph talks about expanding the peephole rewriter, but that got killed in the planning
 
-During the first year of my PhD studies, my research has focused on the
-LeanMLIR project, a framework for defining semantics for MLIR dialects
-within the Lean theorem prover. A central aspect of this work has been
-modelling MLIR dialects with completely pure semantics, exploiting the
-higher-level nature of MLIR dialect, where referential transparency is
-encouraged.
+% The optimization pass will be instantiated with a set of \emph{peephole
+% optimizations}, consisting of program patterns to match for and
+% corresponding program patterns to rewrite to. Lean-MLIR
+% (\cite{bhatVerifyingPeepholeRewriting2024}), a project in
+% development at my research group which I've contributed to, currently
+% has a prototype of such an optimizer. However, the current version is
+% only able to apply \emph{pure} optimizations. I wish to expand this to
+% an optimizer that can apply optimizations involving side effects such as
+% memory loads/stores, undefined behaviour, or potentially non-terminating
+% control flow, and, crucially, provide push-button proof automation for
+% such side-effecting optimizations, using symbolic execution.
 
-We need the full power of a theorem prover to be able to capture the
-extensible and high-level nature of MLIR dialects.
-Still, to recover the ease-of-use that SMT-solver-based tools provide, 
-we've also been investing in building solid automation for bitvector proofs in collaboration with the Lean community.
+After optimization, the next phase will be to generate assembly from our
+optimized IR and, crucially, to verify that the generated assembly has
+the same semantics as the optimized IR. For maximum trust, we'd like to
+use authoritative models of the ISA semantics of whichever platform we
+are compiling for to give semantics to the assembly program. Sail
+(\cite{armstrongISASemanticsARMv8a2019}) provides such models: the
+Sail model for Arm-A is automatically derived from the official ASL
+reference, and their RISC-V model has been adopted by the RISC-V
+Foundation. We'll phrase the code generator similarly: by having
+fragments of compiler IR to match for, and corresponding fragments of
+assembly to rewrite them with. Still, code generation presents some
+unique challenges. Firstly, although the transformations themselves
+should be simpler, the Sail semantic model will be much larger than our
+handwritten LLVM semantics. Secondly, we have to deal with semantics of
+different languages (the compiler IR and the assembly language). It is
+likely that proof automation for code generation will thus need a
+slightly different design than proof automation for peephole
+optimizations.
 
-In the next two years, I'd like to look into a feature of MLIR called
-\emph{Graph Regions}. Specific dialects may use such regions to allow a
-variable to be used \emph{before} it is defined --- hence, turning the
-def-use chain into a general graph. Generally, the semantics of such
-dialects are stateful, and cycles are broken by looking at the value of
-a variable at a previous state.
-For example, graph regions are used by
-dialects that model hardware,
-where variables model wires and cycles represent physical cycles in the
-circuit~\cite{eldridgeMLIRHardwareCompiler}.
+Finally, to model the semantics of LLVM IR in a modular way, we'd like
+to follow Zakowski et
+al.~(\cite{zakowskiModularCompositionalExecutable2021}) and use
+interaction trees (\cite{xiaInteractionTreesRepresenting2020}).
+Unfortunately, interaction trees are \emph{coinductive} in nature, to
+capture possibly non-terminating programs, and Lean currently lacks
+support for coinductive type. Therefore, I'd like to continue the work
+that I started in my own master's thesis~\cite{keizerImplementingDefinitionalCodatatype}, which is based on earlier
+work by Avigad et al.~\cite{avigadDataTypesQuotients2019}, and
+continue development of a Lean library for coinductive types.
 
-Since the exact semantics of graph regions may differ between dialects,
-we need some general framework to reason about stateful behaviours in
-Lean. I propose that coinductive types are this framework, and would
-like to continue my earlier work on supporting coinductive reasoning in
-Lean, with the aim of using them to model graph regions in LeanMLIR.
+\subsubsection*{Research Questions}
 
-In fact, coinductive types are useful beyond graph regions. Since
-coinduction models infinite behaviours, it is also a good way to model
-non-termination without giving up on computability of a dialect's
-semantics, thus addressing a current short-coming of LeanMLIR.
+My research will aim to answer the following questions:
+\begin{enumerate}
+  \item[\textsc{RQ1}]\label{research-question-1}
+    How do we fully automatically verify optimizations involving memory side effects on a mid-level compiler IR, 
+    while retaining a minimal trusted code base?
 
-\section{Related work}\label{related-work}
+  \item[\textsc{RQ2}]
+    How do we fully automatically verify lowerings from a mid-level compiler IR
+  to a low-level assembly language, using large-scale, authoritative ISA semantics models?
 
-First, we'll review the literature on verified compilation, which
-naturally leads us to the literature on coinductive types.
+  \item[\textsc{RQ3}] 
+    How do we recover the modularity that interaction trees provide,
+      in a theorem prover that does not support coinduction?
+\end{enumerate}
 
-\subsection{Verified Compilation}\label{verified-compilation}
+\subsubsection*{Criteria For Success}
 
-CompCert is the
-seminal work when it comes to verified compilation: an end-to-end
-verified to not miscompile compiler for the C language~\cite{leroyCompCertFormallyVerified}.
-However, exactly
-because it's an end-to-end verified monolith, CompCert is not
-competitive compared to traditional non-verified compilers. Outside
-certain high-risk domains, where traditionally optimizations would be
-turned off entirely~\cite{kastnerCompCertPracticalExperience},
-the vast majority of C code is thus still compiled with non-verified
-compilers, which studies have shown to be prone to miscompilations~\cite{yangFindingUnderstandingBugs2011}.
+Question \textsc{RQ1} can be rephrased as: 
+how do we reproduce Alive2 in a way that does not trust an SMT solver. 
+Luckily, Alive2 has a pretty comprehensive test suite.
+I'll evaluate my tool against this same test suite; it should pass all tests,
+besides those that use unstructured control flow.
 
-Lopes\etal{} propose \emph{translation validation} as a way to scale
-formal verification~\cite{lopesAlive2BoundedTranslation2021}.
-Their implementation, called Alive, has found wide adoption in the
-development of LLVM. Given a declarative description of an optimization
-as a source and target pattern in LLVM IR, the tool uses an SMT solver
-to automatically determine whether the optimized pattern exhibits only
-behaviours of the original program.
+For \textsc{RQ2}, the set of all desired rewrites is known up front: we simply 
+need a lowering for each instruction of the mid-level IR.
+We'll evaluate our instruction selector by building such a complete set of lowerings.
+Ideally, there should be no proof obligation that is specfic to a particular lowering,
+this should all be taken care of automatically.
 
-Alive is able to leverage SMT solvers because the semantics of LLVM IR
-are given in terms of bitvectors, and SMT solvers understand the
-bitvector domain well. However, the semantics of more high-level
-compiler IRs, for which MLIR was created, can involve complex
-mathematical objects, which today's SMT solvers are not able to prove
-properties about. For example, the semantics for the ``Poly'' dialect
-\footnote{https://homomorphicencryption.org/} are generally given in
-terms of finitely-presented commutative rings. Furthermore, MLIR is, by
-design, extensible, so a translation validation tool for MLIR also needs
-to be easily extensible to cover new dialects.
+Finally, we ask \textsc{RQ3} because we need interaction trees to define the semantics
+of potentially infinite loops. 
+We'll consider our interaction tree library successful if
+we can provide a similar basic API, and in particular, if we can define such semantics
+without referring to details of the underlying QPF construction directly.
 
-Our group has been developing
-LeanMLIR, a framework for giving
-semantics to MLIR-style compiler IRs in the Lean theorem prover~\cite{demouraLeanTheoremProver2015}. 
-As a theorem prover, Lean is
-much more expressible than SMT solvers. Among theorem provers, Lean is
-particularly suited for this purpose because of its community of
-mathematicians developing a comprehensive mathematical library~\cite{themathlibcommunityLeanMathematicalLibrary2020},
-and because of its highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.
+\subsubsection*{Verified Compilation and Lean's Extensibility}
 
-However, using a theorem prover means we don't get as much automation
-out of the box, and one area where Lean is still lacking is in its lack
-of support for coinductive types.
+Note that the benefits of a verified compilation pipeline go far beyond
+simply reducing the chance of bugs. We believe it will lead to a culture
+shift in the way software is written. Traditionally, writing compiler
+optimizations was a domain reserved for the wise elders who we could
+trust to perform the appropriate ``black magic''. Reasoning about the
+correctness of a transformation was simply too hard, so we had to rely
+on interpersonal trust. Tools like Alive have already caused a shift in
+this approach: nowadays PRs that propose adding a new peephole
+optimization to LLVM can include a link to Alive's online interface
+showing the correctness, meaning no further discussion or review is
+required. However, this effect is still limited to making reviews for
+proposed additions to LLVM easier.
 
-\subsection{Coinductive Types}\label{coinductive-types}
+Lean, again, presents a very interesting opportunity because of it's
+\emph{extensibilty}: Lean already has an extensible syntax and a
+powerful meta-programming API. In fact, a lot of features that are
+considered part of Lean's core are actually implemented using the same
+meta-programming facilities that are available to libraries (we'll use
+this extensively in~\autoref{a-codatatypes-library-for-lean}). However, we imagine a future where this
+extensibility could extend to the backend of the compiler also, such
+that a library writer might implement an algorithm in a functional style
+that is easy to reason about, but also provide an optimized assembly
+\emph{and} a proof that the assembly is equivalent. Traditionally,
+hard-coded assembly in a library has been cause for increased scrutiny.
+Optimizing is hard, do we really trust the author of this library to
+have done it correctly? Verification means we no longer have to trust,
+instead we get to have fearless bespoke optimizations!
 
-Lean, like most theorem provers, is built around a native concept
-of inductive types. This is no surprise, given that most mathematical
-constructions are inductive in nature. Coinductive constructions, in
-contrast, are less common in general mathematics, but, they are crucial
-in computer science.
+\section{High-Assurance Translation Validation for Compiler
+Optimizations}\label{high-assurance-translation-validation-for-compiler-optimizations}
 
-For us, the main difference of interest is that inductive datastructures
-are finite, while coinductive datastructures capture infinite
-behaviours~\cite{sangiorgiIntroductionBisimulationCoinduction2011}.
-In particular, since Lean requires us to prove that recursive functions are
-terminating, we need corecursion to correctly model things like infinite
-loops and other non-terminating behaviours that are common in compiler
-IRs.
+To answer research question \textsc{RQ1}, I'll build proof automation, in Lean,
+which will decide equivalence of program fragments in a compiler IR based on LLVM IR.
 
-For example, Xia\etal{} used interaction trees, a coinductive type, in
-their Coq formalization of recursive and impure programs~\cite{xiaInteractionTreesRepresenting2020}.
-Similarly, MLIR has an
-``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
-centred on finite state machines. Such automata are understood to be
-coinductive in nature, so mechanizing the semantics of this dialect in
-Lean will likely require coinduction.
+LLVM IR is an intermediate representation that sits just above the level
+of abstraction of assembly, and has proven a successful platform to
+perform optimizations on before finally lowering to platform-specific
+assembly. LLVM has been used as a back-end for many mainstream
+compilers, such as Clang, Rust and Swift. A crucial property of LLVM IR, is that it is in static single assignment
+(SSA) form: rather than operating on a fixed set of registers, LLVM IR
+has variables that are assigned exactly once. This form makes it easier
+to apply peephole rewrites, as we can simply match for a pattern along
+what is called the def-use chain, even if there are unrelated
+instructions in between.
 
-Consequently, we need a user-friendly way to define coinductive
-structures and to formalize coinductive proof methods in Lean. Key prior
-work in this area includes my own master's thesis (\cite{keizerImplementingDefinitionalCo}), 
-which is based on earlier
-work by Avigad\etal{} that models coinductive as quotients of polynomial
-functors (QPFs)~\cite{avigadDataTypesQuotients2019}.
+Now, although LLVM is a bit higher level than assembly, it's still
+low-level and, e.g., features unstructured control flow. The LLVM
+community realized that having more structure available would enable
+even more optimizations to be done purely locally (i.e., as peephole
+rewrites). Thus, MLIR (Multi Level Intermediate Representation) was
+born: a framework to define compiler IRs at various levels of
+abstraction. MLIR mandates that IRs are in SSA form, and takes care of
+syntactic overhead. We can instantiate the framework with a specific set
+of operations, called a dialect. Generally, a compiler IR is made up of
+a combination of such dialects. For this chapter, we shall focus on an
+IR that contains bitvector arithmetic (a fragment of the \texttt{arith}
+dialect) memory manipulations (\texttt{ptr} dialect) and structured
+control flow (i.e., for and while loops, \texttt{scf} dialect). 
+The arithmetic and pointer operations of these dialects have exactly the same semantics as in LLVM, 
+but using \emph{structured} control flow puts us just above the level of abstraction of LLVM IR.
 
-Crucially, this prior work builds coinductive types using constructions
-already supported by Lean today. This allows us to provide coinductive
-types as a \emph{library}, instead of having to modify the logical
-system itself to support coinduction natively --- an approach that is
-greatly aided by Lean's excellent meta-programming capabilities
-mentioned previously.
+\subsection{Lean-MLIR Today}\label{lean-mlir-today}
 
-When looking at coinductive types in other theorem provers, we
-distinguish between languages like Isabelle~\cite{traytelCategoryTheoryBased}, which follow the same
-coinduction-as-a-library approach, and languages like Coq~\cite{gimenezTutorialRecursiveTypes1998,gimenezApplicationCoinductiveTypes1996}
+Lean-mlir is a framework for modelling the \emph{semantics} of MLIR
+dialects in Lean, developed by my research group~\cite{bhatVerifyingPeepholeRewriting2024}, focusing on the
+\emph{pure} (i.e., side-effect-free) fragment of MLIR dialects.
+
+In particular, lean-mlir currently has a model of just the bitvector
+arithmetic operations of LLVM IR, and a verified peephole rewriter that
+can apply a pure peephole rewrite (i.e., neither the program fragment
+being matched on nor the fragment being rewritten to contain
+side-effecting operations) and, assuming the local correctness of every
+rewrite, we've shown the rewriter preserves the global semantics.
+
+Furthermore, the Lean FRO has, in collaboration with my research group
+and me, been developing proof automation for deciding the equality
+of fixed-width bitvector expressions\cite{bovingTamingBitvectorBestiary}. In
+particular, this proof automation allows us to fully automatically
+decide the local correctness of \emph{pure} optimizations. Note that
+although this tactic employs an external (certificate-producing) SAT
+solver, it also includes a verified certificate checker, so that the
+final proofs do \emph{not} rely on the correctness of the solver. That
+said, the checker would be too slow if ran inside the Lean kernel, so to
+get decent performance, this tactic \emph{does} rely on the previously
+mentioned feature to run the compiled version of a Lean tactic, thus expanding
+the trusted code base to include the Lean compiler.
+As the paper argues, this is still a much smaller trusted code base than a full
+SMT solver.
+
+\subsection{Proof Automation for Side
+Effects}\label{proof-automation-for-side-effects}
+
+My contributions to lean-mlir are primarily focussed on expanding the framework to 
+include support for impure operations with side effects. 
+I wish to continue this effort by modelling all operations of the
+dialects mentioned before, in particular the dialect dealing with memory.
+The bitvector arithmetic fragment of the previous section is basically the same as \texttt{arith}, and the
+framework also already models part of \texttt{scf} (in particular, it
+models counting for-loops, but not while-loops with arbitrary conditions
+yet), so the main missing piece is a model of memory and memory
+operations.
+
+The main innovation in this project will be to build proof automation in
+Lean for translation validation on this compiler IR. That is, given a
+source and target program (potentially including holes), the tool will
+automatically decide the correctness of the transformation. To do this
+in the presence of side effects, I'll build a symbolic simulator.
+
+\subsection{Refinement and Undefined
+Behaviour}\label{refinement-and-undefined-behaviour}
+
+So far, we've talked about semantic equivalence as the condition for
+correctness of an optimization, but in reality it's a bit more
+complicated because of the notion of undefined behaviour (UB). In short,
+if a program has UB, it's legal to change the program to anything.
+Conversely, if a source program is fully defined, then it's illegal to
+introduce UB in the target. The corresponding relation between programs
+is called \emph{refinement}.
+
+\subsection{Related Work}
+
+We can broadly categorize earlier efforts to apply formal methods to
+LLVM into those that formalize the semantics of LLVM, but perform manual
+proofs (e.g., VeLLVM~\cite{zhaoFormalVerificationSSABased, zhaoFormalizingLLVMIntermediate2012} 
+or work by Zakowski et al.~\cite{zakowskiModularCompositionalExecutable2021}) and tools
+like Alive (\cite{lopesAlive2BoundedTranslation2021}) that perform
+fully automated translation validation, but rely on SMT solvers.
+
+AliveInLean is a prototype reimplementation of Alive in Lean that shows how to increase trust without
+sacrificing automation~\cite{leeAliveInLeanVerifiedLLVM2019}. AliveInLean helps to justify the reduction from a model of LLVM semantics
+to SMT queries. Nonetheless, AliveInLean still trusts the SMT solver and does not support reasoning about memory. 
+Our tool should address both of these concerns.
+
+More broadly, CompCert~\cite{leroyCompCertFormallyVerified} and
+CakeML~\cite{kumarCakeMLVerifiedImplementation2014} are examples
+of fully end-to-end verified compilers. More recent work by Mullen et
+al.~\cite{mullenVerifiedPeepholeOptimizations2016} argues for
+peephole rewriting as a way to lower proof burden in CompCert
+development, but they still prove these rewrites manually.
+
+Katamaran \cite{keuchelVerifiedSymbolicExecution2022} is a
+symbolic simulation tool written in Coq. However, this
+project focusses on being able to run the simulator outside the ITP
+environment, hence eschewing meta-programming. Our main purpose is
+modular proof automation, so being tied to Lean is no problem for us,
+allowing us to make different implementation choices.
+
+\section{Scaling High-Assurance Translation Validation to
+Authoritative ISA
+Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models}
+
+After performing optimizations on the level of our compiler IR, we need
+to lower this representation to assembly. We'll do this in a very
+similar way, by phrasing instruction selection as rewrites that match
+for a compiler IR program fragment and rewrite it to an assembly program
+fragment. By following a similar approach, we can hopefully reuse some
+of the infrastructure we developed in the previous chapter.
+
+By building proof automation in Lean that can automatically verify the correctness of
+such lowering, I'll answer research question \textsc{RQ2}.
+
+\subsection{The RISC-V Dialect}\label{the-risc-v-dialect}
+
+Lopoukhine et al have developed a set of MLIR
+dialects that together model RISC-V assembly~\cite{lopoukhineMultiLevelCompilerBackend}. Representing our target
+for code generation as an MLIR dialect allows us to use Lean-MLIR,
+enabling reuse of the core datastructures provided by the framework, but
+the benefits extend much further! Namely, depending on which of the
+dialects provided by Lopoukhine et al.~we use, our representation of
+assembly can (a) be either register allocated, or as an assembly-like
+SSA IR, and (b) use either structured (for/while loops) or unstructured
+control flow (branches).
+
+We'll use these progressively lower level IRs as targets for our code
+generator, explicitly separating the instruction selection pass, which
+we'll implement via a simple rewriter, from a register allocation pass.
+Finally, recall that we chose to stick to structured control flow in the
+optimization pass to preserve structure. We'll initially do the same in the code
+generator, postponing the lowering from structured into unstructured
+control flow to the last moment.
+
+\subsection{Semantics for RISC-V}
+
+So far, we've eliminated any reliance on SMT solvers in pursuit of a
+smaller trusted code base. But recall that the semantics model of
+whichever language we choose to prove things about is also a crucial
+part of the trust story. In the end, if we run our compiler on actual
+hardware, then a proof of correctness is only good so long as the actual
+hardware behaves the same as our model of the instruction set
+architecture (ISA).
+
+One could attempt to formally verify the hardware as well, but that is
+well beyond the scope of my PhD. Instead, I'd like to use Sail models as
+an authoritative source of truth for ISA semantics~\cite{armstrongISASemanticsARMv8a2019}.
+Sail is a domain specific
+language for modelling ISA semantics. Various models exist, but in
+particular, the Arm-v8 Sail model has been automatically derived from the Arm-internal 
+specification~\cite{reidTrustworthySpecificationsARM} and the RISC-V model has
+been adopted by RISC-V International~\cite{sammlerIslarisVerificationMachine2022}. 
+Using these models as the source of our semantics gives us trust that any discrepancy
+between our formalization and actual hardware likely is a bug in the
+hardware!
+
+The Sail project currently provides a translation from Sail to Coq,
+which has been used previously to formally reason about ISAs. Luckily, a
+translation from Sail to Lean is currently being developed, following a
+similar strategy to add a backend to sail that emits Lean code---as
+opposed to embedding a model of Sail in Lean.
+
+Of course, the Sail semantics will operate on traditional assembly, with
+registers allocated and unstructured branches. This should not be a
+problem, as lowering a single instruction at any of the progressive
+levels of abstraction down to regular assembly should not be difficult.
+
+\subsection{Translation Validation}\label{translation-validation}
+
+Again, the core contribution will be proof automation. By treating
+instruction selection as a sequence of rewrites, we merely need to
+verify that each of these rewrites individually preserves semantics.
+
+This seems like much the same problem that we tackled in the previous
+chapters, but there are some subtleties. For one, the rewrites
+themselves should be simpler, with many rewrites having only a single
+instruction as source and target. However, a Sail-derived semantics
+model will be much larger than the handwritten model we'll be using for
+the compiler IR, necessitating many more reasoning steps to simplify the
+semantics for a single instruction. Many of the details that make the
+Sail model so large might be irrelevant in our context, so I'll
+investigate ways to simplify the model under specific assumptions. For
+example, by assuming we're only compiling user-level programs, we might
+be able to simplify away any complexity to do with kernel-level
+instructions in the model.
+
+\subsection{LNSym}\label{lnsym}
+
+During my internship at Amazon, I developed a symbolic simulator for a
+simplified fragment of Arm assembly programs, using a handwritten model
+of the ARM ISA semantics. My internship focused primarily on making this
+symbolic simulator scale to being able to simulate hundreds of
+instructions within a reasonable timeframe. I'll be using this symbolic
+simulator as a basis for the translation validation tool I aim to
+develop, adapting it to the slightly different design constraints in
+this context.
+
+\subsection{Related Work}\label{related-work}
+
+Isla is a symbolic simulator for Sail, which takes in a program in assembly 
+together with a Sail model for that assembly language, and outputs a trace of the
+execution in Isla Trace Language~\cite{armstrongIslaIntegratingFullScale2021}.
+However, Isla relies on an SMT solver to produce this trace.
+
+Islaris is a Coq development, which builds upon Isla and Iris separation logic to 
+prove properties about assembly programs using the power and expressivity of a 
+theorem prover~\cite{sammlerIslarisVerificationMachine2022}. Their strategy starts by using Isla to transform a program into a trace, and then proving properties about this trace.
+To remove the need to trust the SMT solver from, Sammler et al also investigated
+the use of automated translation validation, which certifies that the trace produced
+by Isla agrees with the lowering of the Sail semantics to Coq. This approach 
+worked well for RISC-V, but unfortunately did not scale to the much larger Arm model.
+
+\section{A Codatatypes Library for
+Lean}\label{a-codatatypes-library-for-lean}
+
+In previous chapters, we've glossed over how, exactly, I'll give
+semantics to the side-effectful version of our compiler IR. Zakowski et
+al.~have written about the benefits of an interaction tree (ITree)
+based semantics~\cite{zakowskiModularCompositionalExecutable2021, xiaInteractionTreesRepresenting2020}, so I'd like to use
+ITrees also. Sadly, ITrees are coinductive, to properly capture infinite loops, 
+and Lean does not support coinductive types. 
+I'd like to address this, and build an ergonomic Lean library for ITrees,
+answering research question \textsc{RQ3} in the process.
+
+\subsection{Quotients of Polynomial
+Functors}\label{quotients-of-polynomial-functors}
+
+Luckily, we can encode coinductives in Lean's existing logic, using
+so-called \emph{quotients of polynomial functors} (QPFs). Avigad et
+al.~have written about the low-level construction~\cite{avigadDataTypesQuotients2019}, 
+upon which I expanded in my Master's thesis by prototyping a higher-level framework for defining
+coinductive types, called QPFTypes. This framework is far from perfect,
+and I'll spend some time fixing bugs in the framework. Still, the main
+goal will be simply to define interaction trees; not to perfectly polish
+the framework.
+
+\subsection{Corecursive Functions}\label{corecursive-functions}
+
+The QPFTypes framework currently only has facilities for defining types;
+to define corecursive functions that operate over ITrees we only have
+the low-level corecursion principle to work with. However, in their
+original development of ITrees, Xia et al. in fact provide a set of three
+combinators which they claim make working with ITrees more tractable,
+and remove the need for users to work directly with Coq's implementation
+of corecursive function (which they claim is not compositional). As it
+turns out, these combinators are rather similar to the corecursion
+principle we do have available. So, for our ITree library
+in Lean, we don't have to worry about general corecursive functions; it
+suffices to implement the three basic combinators that Xia et
+al.~provide.
+
+\subsection{Related Work}\label{related-work-1}
+
+The QPF construction encodes coinductive types as a \emph{library},
+instead of having to modify the logical system of Lean to support
+coinduction natively --- an approach that is greatly aided by Lean's
+excellent meta-programming capabilities. When looking at coinductive
+types in other theorem provers, we distinguish between languages like
+Isabelle~\cite{traytelCategoryTheoryBased}, which follow the same
+coinduction-as-a-library approach, and languages like Coq~\cite{gimenezTutorialRecursiveTypes1998, gimenezApplicationCoinductiveTypes1996}
 or Agda which have modified their trusted kernels to support
@@
 coinductive types, in terms of \emph{bounded natural functors} is
-closely related to our construction in terms of QPFs~\cite{furerQuotientsBoundedNatural2022}. 
-There are still new
+closely related to our construction in terms of QPFs~\cite{furerQuotientsBoundedNatural2022}. There are still new
 challenges, however, because Lean's dependent types mean we have to
@@
 
-In contrast, the coinduction-in-the-kernel approach allows us to modify
-the kernel to understand coinduction and ensure the kernel recognizes
-the desired definitional equalities. However, modifying the trusted
-kernel carries a large burden of proof, since the changes could
-compromise the logical soundness of the entire system. In fact, the
-original implementation of coinduction, called \emph{positive}
-coinduction, is nowadays discouraged in favour of an alternative
-implementation, since positive coinduction breaks the subject reduction
-property\footnote{The subject reduction property states that reducing a program does not change its type}~\cite{sozeauCorrectCompleteType}.
-Our coinduction-as-a-library approach requires no new axioms nor
-changes to the kernel, and is thus guaranteed not to change any meta
-properties of Lean.
+In contrast, the coinduction-in-the-kernel approach would modify the
+kernel to understand coinduction and ensure the kernel recognizes the
+desired definitional equalities. However, modifying the trusted kernel
+carries a large burden of proof, since the changes could compromise the
+logical soundness of the entire system. In fact, the original
+implementation of coinduction, called \emph{positive} coinduction, is
+nowadays discouraged in favour of an alternative implementation, since
+positive coinduction breaks the subject reduction property
+\footnote{The subject reduction property states that reducing a program
+  does not change its type}\cite{sozeauCorrectCompleteType}. The
+coinductives-as-a-library approach requires no new axioms nor changes to
+the kernel, and is thus guaranteed not to change any meta properties of
+Lean.
 
-\section{Progress Report}\label{progress-report}
+\section{Risks and Mitigations}
 
-The bulk of my time has been spent developing the LeanMLIR framework, 
-in collaboration with other members of our research group, 
-culminating in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
-Compiler IRs''}.
+A mayor risk is that my proposed plan relies on a translation from Sail to Lean 
+which is being worked on, but does not yet exist today. Luckily, this translation
+is following the footprints of the existing Sail to Coq translation, 
+and Lean and Coq are similar enough that there is no reason to expect mayor difficulties.
+Furthermore, the SailToLean project currently has a lot of interest.
+Still, to allow as much time as possible, I've pushed back work on the chapter that needs this
+translation to the end of my PhD. If the translation still does not work as needed by that time,
+I can always fall back to a handwritten ISA semantics, as we used in LNSym.
 
-\subsection{Lean MLIR}\label{lean-mlir}
+Another risk is that Lean is a relatively less mature theorem prover.
+In particular, the style of program verification proofs that will be generated by the proof automation
+I propose might be rather different from the type of proofs that most users of Lean
+today generate; and thus we might encounter some performance problems inherent to Lean.
+In fact, we encountered, and reported, exactly such performance problems while working on LNSym, and this experience means I now know how to debug and mitigate similar
+problems.
+If my proposed automation does run into limitations of Lean itself,
+I'll reduce the scope of what I aim to prove automatically, and 
+evaluate and report exactly where the bottlenecks lie.
 
-LeanMLIR is a framework for giving semantics to MLIR dialects in Lean.
-On top of this framework, we've built a peephole rewriter, and proved
-that if a rewrite locally preserves semantics, then applying this
-rewrite in the context of a larger program will preserve the semantics
-of the program as a whole.
+\section{Time plan}\label{timeplan}
 
-However, we've focused on dialects with pure (i.e., side effect free)
-semantics. Even though we \emph{are} able to verify a lot of interesting
-rewrites in high-level MLIR dialects, we do need to be able to reason
-about side effect if we want to capture the low-level dialects as well,
-as a reviewer of the paper rightly pointed out.
+My PhD clock started in Michealmas term 2023, meaning my submission
+window is between October 2026 and September 2027, or between 21 and 33
+months from now. Picking the lower end of that timeframe gives exactly 7
+months per chapter, leaving the final year for writing up the thesis, which I'll
+spend as follows.
 
-To this end, we've generalized the framework so that a dialect's
-semantics may be given in a user-defined monad, and proven that applying
-a semantics-preserving pure rewrite (i.e., where the operations being
-rewritten are all side effect free) inside a side-effecting program
-preserves semantics of the program as a whole. However, this
-generalization still has a lot of rough edges, and further work is
-needed before we can claim our framework is versatile enough to deal
-with side effects cleanly.
+\subsection*{High-Assurance Translation Validation for Compiler Optimizations
+Targeted publication venue: PLDI}
 
-\subsection{Bitvectors}\label{bitvectors}
+\begin{itemize}
+\item
+  Translation Validation for arithmetic \& bounded, structured control
+  flow (2 months)
+\item
+  Translation Validation for memory side effects (2 months)
+\item
+  Stress-test and improve scalability of proof automation (2 months)
+\item
+  Paper writing (1 month)
+\end{itemize}
 
-As mentioned, a big reason for us to use Lean is its community. As
-such, I've been heavily involved in the redesign of Lean's API for
-bitvectors, and the move of bitvectors from Mathlib (the mathematical
-library previously mentioned) to become part of Lean's core library
-(i.e., the library that is by default included in every Lean program).
-By ensuring our representation of bitvectors is aligned with the rest of
-the community, we ensure we can easily reap the benefits of, e.g.,
-tactics to automate bitvector proofs made by others.
+\subsection*{A Codatatypes Library for
+Lean}\label{a-codatatypes-library-for-lean-1}
 
-One such automation effort is the LeanSAT\footnote{https://github.com/leanprover/leansat}
-library, which aims to automate proofs about bitvectors of bounded
-widths, by implementing a verified bitblaster and SAT sovler in Lean.
-Siddharth (another PhD student in our group) and I have been
-contributing to this project by formally proving necessary facts in the
-core library.
+Targeted publication venue: ITP
 
-This automation is crucial, 
-because it brings the user experience for non-experts to the same level as Alive, 
-for dialects whose semantics use bitvectors, 
-while still having the full expressivity needed for dialects that use more complex mathematical
-objects. Thus, addressing the main drawback of using a theorem prover.
+\begin{itemize}
+\item
+  Bugfix framework to accept ITree type definition (2 months)
+\item
+  Define basic ITree API, including combinators (2 months)
+\item
+  Evaluate ITrees by expanding compiler IR semantics to infinite loops
+  (2 months)
+\item
+  Paper writing (1 month)
+\end{itemize}
 
-\section{Future plans}\label{future-plans}
+\subsection*{Scaling High-Assurance Translation Validation to
+Authoritative ISA
+Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models-1}
 
-\subsection{Bitvectors}\label{bitvectors-1}
+Targeted publication venue: PLDI
 
-\begin{quote}
-June '24 - November '24
-\end{quote}
-
-As mentioned in the progress report, I'm currently contributing to a
-verified bitblasting procedure in Lean, called LeanSAT.
-
-\subsubsection{Arbitrary widths}\label{arbitrary-widths}
-
-LeanSAT is, however, not a silver bullet. Just like SMT solvers, it can
-only reason about bitvector expressions where the width of the
-bitvectors involved is bounded --- and even then, only if the bound is not too large.
-One of the topics I'd like to explore more is the connection between bitvectors and 2-adic numbers (which are
-infinite and thus coinductive in nature!) to build a decision procedure
-which can decide equality of a certain (limited) class of
-arbitrary-width bitvector expressions.
-
-Furthermore, during my earlier work on LeanMLIR, we've already encountered a 
-few other strategies for solving other classes of arbitrary-width bitvector problems.
-So, even though deciding these problems in general is undecidable, we can identify
-specific classes of problems that do have decision procedures.
-
-I'm planning to write a paper about this work tentatively titled: ``\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}'', 
-in which we explore all of these different strategies, 
-classify exactly which class of expressions they solve, 
-and investigate how many real-world problems are already covered by these strategies.
-The aim is to submit this to PLDI, which has its call for papers around November.
-
-\subsubsection{Amazon Internship}\label{amazon-internship}
-
-\begin{quote}
-July '24 - November '24
-\end{quote}
-
-Furthermore, I plan to continue this work on bitvector automation in
-Lean during my internship at Amazon\footnote{Assuming I'll get
-  permission to work away, of course}. They are particularly interested
-in automatically reasoning about ARM assembly code, which closely aligns
-with our goal of automating bitvector reasoning to then use for
-verifying (low-level) MLIR dialects.
-
-This internship will thus provide me with exactly the corpus of real-world bitvector verification problems that I need to evaluate the strategies we'll discuss in the 
-aforementioned paper.
-
-\subsection{Coinduction}\label{coinduction}
-
-The next big theme to tackle will be to make the Lean library for
-coinductive types ready for serious use, which mainly involves
-high-level support for corecursive functions.
-
-\subsubsection{Productivity Checking}\label{productivity-checking}
-
-\begin{quote}
-June '24 - Aug '24 (Supervising Internship)
-\end{quote}
-
-Currently, the \texttt{QpfTypes} framework is able to compile a
-high-level description of a coinductive type into a construction in
-terms of concepts Lean already supports. As part of this construction, a
-\emph{corecursion principle} is generated. For example, the corecursion
-principle for \texttt{Stream\ \textbackslash{}alpha}, the type of
-infinite lists (i.e., streams) over elements of type
-\texttt{\textbackslash{}alpha}, is as follows:
-
-\begin{verbatim}
-Stream.corec : (\beta -> \alpha \times \beta) -> \beta -> Stream \a
-\end{verbatim}
-
-While this corecursion principle is in theory all that's needed to
-define functions over streams, it's not directly usable in practice.
-Hence, the first subgoal will be to support a high-level way to define
-corecursive functions, following the idiomatic way of defining inductive
-functions in terms of pattern matching, rather than recursion
-principles, which is then compiled into an application of this
-corecursion principle. This will involve some software engineering,
-dealing with Lean's extensible syntax, plus a good amount of research,
-to know which functions are \emph{productive} in our formalization, and
-to automatically recognize (a well-defined subclass of) these
-productive corecursive functions.
-
-Productivity in this context is the dual of termination in recursive
-contexts; a corecursive function is said to be productive if, at all
-times, it will produce the next element of our infinite corecursive
-structure in finite time. Ensuring all corecursive functions are
-productive is paramount to our logical system remaining consistent. In
-fact, because of our coinduction-as-a-library approach we are unable to
-define corecursive functions without proving they are productive, and
-conversely, if we can get the system to accept a corecursive function,
-we know that allowing this function must be productive, and allowing it
-is sound.
-
-I will be a co-supervising an undergraduate student over the summer who
-will be implementing support for corecursive functions using a
-\emph{guarded} productivity checker. That is, this first implementation
-will only allow functions where each corecursive call occurs behind
-a constructor. This check is known to incomplete, and more complicated
-schemes have been studied
-(\cite{blanchetteFriendsBenefitsImplementing2017}), but a
-guardedness easier to implement and will most likely be sufficient for
-now --- the initial implementation of corecursion in Coq also allowed
-only guarded corecursion~\cite{gimenezTutorialRecursiveTypes1998}.
-Of course, if there is still time left in the internship after this goal
-has been reached, we will attempt to implement more complicated schemes
-as well.
-
-\subsubsection{Graph Regions \& Parametric
-Coinduction}\label{graph-regions-parametric-coinduction}
-
-\begin{quote}
-December '24 - June '25
-\end{quote}
-
-A \emph{Graph Region} in MLIR is, just like a regular region, a sequence
-of instructions with their results bound to variables. However, unlike
-normal regions, in a graph region the \emph{use} of a variable may
-precede its definition. For example
-
-\begin{verbatim}
-^bb(%x : i32)
-  %y = comb.and %x %y : (i32, i32) -> (i32)
-\end{verbatim}
-
-Here, \texttt{y} is obtained as the bitwise and of the argument
-\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
-sense, but this style of definition is common in hardware
-representations, where this is interpreted as: ``the value of \texttt{y}
-in the \emph{current} clock cycle is computed by taking the bitwise and
-of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles.
-See https://circt.llvm.org/ for more details}.
-To formally model this, we could denote variables as infinite streams of bitvectors,
-which gives us the value at every possible future clock cycle. Another
-example of a dialect with graph regions is the FSM, or Finite State
-Machine, dialect.
-
-Graph regions are thus inherently stateful, and modelling them is a
-particularly good use-case for coinductive types and coinductive
-reasoning. To facilitate easy verification of coinductive properties in
-this setting, I would like to investigate supporting parametrized
-coinductive proofs~\cite{hurPowerParameterizationCoinductive} in
-the coinductive types framework.
-
-Once we've got this, I'd like to actually write a paper about the coinduction library, using this work on graph regions as a case study.
-The goal would be to submit this to POPL 25.
-
-\subsection{LeanMLIR}\label{leanmlir}
-
-Finally, I'd like to explore further uses of coinduction in the formalization of compiler IRs.
-
-\subsubsection{Non-termination}\label{non-termination}
-
-\begin{quote}
-June '25 - September  '25
-\end{quote}
-
-In the status report, we discussed how the framework was generalized to
-work in arbitrary monads. But, to actually define the semantics of an
-MLIR dialect which allows for infinite loops, we need a concrete monad
-that can support this behaviour.
-
-We could, of course, model non-termination using the \texttt{Option}
-monad, where divergence is simply modelled as \texttt{Option.none}. Such
-semantics are necessarily non-computable, losing the very nice property
-that the formalization of a dialect in LeanMLIR also gives a reference
-implementation for an interpreter.
-
-However, coinduction is actually a very good fit for this problem as well.
-Therefore, I'd like to use my library for coinductive types to define a
-coinductive monad which supports computable infinite behaviours, drawing
-inspiration from the interaction tree library for Coq~\cite{xiaInteractionTreesRepresenting2020}.
-This addresses a
-weakness in LeanMLIR while simultaneously giving a nice case study for
-the usability of the coinduction library.
-
-\subsubsection{Runtime behaviour}\label{runtime-behaviour}
-\begin{quote}
-September '25 - Jan '26
-\end{quote}
-
-Once we start using this interaction tree datastructure in LeanMLIR, though, 
-we will likely run into performance problems if we try to actually execute our computable semantics.
-
-A cornerstone of the coinduction-as-a-library approach is that we're
-limited to whatever concepts our theorem prover already supports. The
-QPF construction starts by realizing that an element of an inductive
-type is basically a (finite) tree, while an element of a coinductive
-type is a non-wellfounded (i.e., possibly infinite) tree.
-
-Finite trees are easily encoded in Lean, but representing a finite tree
-is a bit more tricky. The current construction works by representing a
-non-wellfounded tree as a function from a natural number \(n\) to the
-approximation of the non-wellfounded tree up to a maximal depth of
-\(n\), together with a proof that each subsequent approximation agrees
-with the previous approximations up to depth \(n-1\).
-
-This construction is mathematically elegant, but seems to lead to
-unexpected inefficient computational behaviour, as shown by some early
-experimentation by myself. I plan to investigate this more thoroughly,
-but I strongly suspect this behaviour comes from the impedance mismatch
-between the construction of infinite trees as just described, and the
-fact that applications of corecursion principles are essentially state
-machines. I would thus like to explore the possibility of building the
-QPF constructions on a representation of non-wellfounded trees in terms
-of state machines (or, more generally, coalgebras).
-
-% \subsection{Stretch Goal: Iris}\label{stretch-goal-iris}
-% 
-% \begin{quote}
-% {[}!TODO{]} Talk about how denotational semantics are cool, and how the
-% aforementioned improvements will let us define dialects with
-% side-effects, but reasoning about them will still be a bit painful. We'd
-% really like to reason with program logics instead, in particular, iris:
-% \cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
-% \end{quote}
-
-% \input{y1review.tex}
-
-% \begin{figure}
-% % Link to figure
-% %
-% % https://docs.google.com/drawings/d/1juKp43D3rLC-luBQPwQZ_wCnDK2S_6C1k6USV0wKE0g/edit?usp=sharing
-% \includegraphics[width=\columnwidth]{overview.pdf}
-% \caption{Our key idea visualized}
-% \grosser{Replace this figure with your own drawing.}
-% \end{figure}
+\begin{itemize}
+\item
+  Define lowering from compiler IR to assembly (2 months)
+\item
+  Translation validation for unstructured control flow (2 months)
+\item
+  Evaluate and improve scalability of SailToLean-generated ISA model (2
+  months)
+\item
+  Paper writing (1 month)
+\end{itemize}
 

### 2025-01-31 paper.tex
fix publication venue

@@
 months from now. Picking the lower end of that timeframe gives exactly 7
-months per chapter, leaving the final year for writing up the thesis, which I'll
-spend as follows.
+months per chapter, leaving the final year for writing up the thesis.
 
-\subsection*{High-Assurance Translation Validation for Compiler Optimizations
-Targeted publication venue: PLDI}
+Furthermore, I plan to translate each of the technical chapters described in this proposal into a paper, to be submitted at the designated venue.
+
+\subsection*{High-Assurance Translation Validation for Compiler Optimizations}
+Targeted publication venue: PLDI
 
@@
 
-\subsection*{A Codatatypes Library for
-Lean}\label{a-codatatypes-library-for-lean-1}
+\subsection*{A Codatatypes Library for Lean}\label{a-codatatypes-library-for-lean-1}
 

### 2025-01-31 paper.tex
final refinement pass

@@
 algorithm separately from specific transformation instances, and then
-develop proof automation for fully automatically showing local
-correctness of a transformation in a way that composes into a proof of
-global correctness?
+develop proof automation to show local
+correctness of a transformation, without needing input from a formal methods export?
 
-Specifically, I'll develop two core components of a verified compiler:
+I'll investigate two core components of a verified compiler:
 an optimization pass based on peephole optimizations and a code
@@
 \cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
-included no proof automation. This will to our knowledge be the first
+included no proof automation. This will, to my knowledge, be the first
 effort to provide fully automated proofs for correctness of LLVM
@@
 optimized IR and, crucially, to verify that the generated assembly has
-the same semantics as the optimized IR. For maximum trust, we'd like to
+compatible semantics. For maximum trust, we'd like to
 use authoritative models of the ISA semantics of whichever platform we
-are compiling for to give semantics to the assembly program. Sail
-(\cite{armstrongISASemanticsARMv8a2019}) provides such models: the
+are compiling for to give semantics to the assembly program. Sail~\cite{armstrongISASemanticsARMv8a2019} provides such models: the
 Sail model for Arm-A is automatically derived from the official ASL
-reference, and their RISC-V model has been adopted by the RISC-V
+reference, and the RISC-V model has been adopted by the RISC-V
 Foundation. We'll phrase the code generator similarly: by having
@@
 that I started in my own master's thesis~\cite{keizerImplementingDefinitionalCodatatype}, which is based on earlier
-work by Avigad et al.~\cite{avigadDataTypesQuotients2019}, and
+work by Avigad\etal{}~\cite{avigadDataTypesQuotients2019}, and
 continue development of a Lean library for coinductive types.
@@
 
-Finally, we ask \textsc{RQ3} because we need interaction trees to define the semantics
+Finally, I ask \textsc{RQ3} because we need interaction trees to define the semantics
 of potentially infinite loops. 
 We'll consider our interaction tree library successful if
-we can provide a similar basic API, and in particular, if we can define such semantics
-without referring to details of the underlying QPF construction directly.
+we can provide a similar basic API, and in particular, if we can define the 
+semantics we need without referring to details of the underlying QPF construction directly.
 
@@
 Note that the benefits of a verified compilation pipeline go far beyond
-simply reducing the chance of bugs. We believe it will lead to a culture
+simply reducing the chance of bugs. It could lead to a culture
 shift in the way software is written. Traditionally, writing compiler
@@
 
-\subsection{Lean-MLIR Today}\label{lean-mlir-today}
+\subsection{Lean-MLIR Today}\label{Lean-MLIR-today}
 
-Lean-mlir is a framework for modelling the \emph{semantics} of MLIR
+Lean-MLIR is a framework for modelling the \emph{semantics} of MLIR
 dialects in Lean, developed by my research group~\cite{bhatVerifyingPeepholeRewriting2024}, focusing on the
@@
 
-In particular, lean-mlir currently has a model of just the bitvector
+In particular, Lean-MLIR currently has a model of just the bitvector
 arithmetic operations of LLVM IR, and a verified peephole rewriter that
@@
 and me, been developing proof automation for deciding the equality
-of fixed-width bitvector expressions\cite{bovingTamingBitvectorBestiary}. In
+of fixed-width bitvector expressions~\cite{bovingTamingBitvectorBestiary}. In
 particular, this proof automation allows us to fully automatically
@@
 
-My contributions to lean-mlir are primarily focussed on expanding the framework to 
+My contributions to Lean-MLIR are primarily focussed on expanding the framework to 
 include support for impure operations with side effects. 
 I wish to continue this effort by modelling all operations of the
-dialects mentioned before, in particular the dialect dealing with memory.
+dialects mentioned before. In particular, the dialect dealing with memory.
 The bitvector arithmetic fragment of the previous section is basically the same as \texttt{arith}, and the
@@
 automatically decide the correctness of the transformation. To do this
-in the presence of side effects, I'll build a symbolic simulator.
+in the presence of side effects, I'll build a symbolic simulator in Lean, 
+which will rely on the previously mentioned bitvector solver to discharge pure
+bitvector proof obligations.
 
@@
 proofs (e.g., VeLLVM~\cite{zhaoFormalVerificationSSABased, zhaoFormalizingLLVMIntermediate2012} 
-or work by Zakowski et al.~\cite{zakowskiModularCompositionalExecutable2021}) and tools
-like Alive (\cite{lopesAlive2BoundedTranslation2021}) that perform
+or work by Zakowski\etal{}~\cite{zakowskiModularCompositionalExecutable2021}) and tools
+like Alive~\cite{lopesAlive2BoundedTranslation2021} that perform
 fully automated translation validation, but rely on SMT solvers.
@@
 
-\section{Scaling High-Assurance Translation Validation to
-Authoritative ISA
-Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models}
+\section{Automatically Verified Instruction Selection using
+Authoritative ISA Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models}
 
@@
 By building proof automation in Lean that can automatically verify the correctness of
-such lowering, I'll answer research question \textsc{RQ2}.
+such lowerings, I'll answer research question \textsc{RQ2}.
 
@@
 
-Lopoukhine et al have developed a set of MLIR
+Lopoukhine\etal{}have developed a set of MLIR
 dialects that together model RISC-V assembly~\cite{lopoukhineMultiLevelCompilerBackend}. Representing our target
@@
 the benefits extend much further! Namely, depending on which of the
-dialects provided by Lopoukhine et al.~we use, our representation of
-assembly can (a) be either register allocated, or as an assembly-like
+dialects provided by Lopoukhine\etal{}~we use, our representation of
+assembly can (a) be either register allocated, or an assembly-like
 SSA IR, and (b) use either structured (for/while loops) or unstructured
@@
 
-We'll use these progressively lower level IRs as targets for our code
+We'll use these progressively lower level RISC-V IRs as targets for our code
 generator, explicitly separating the instruction selection pass, which
@@
 problem, as lowering a single instruction at any of the progressive
-levels of abstraction down to regular assembly should not be difficult.
+levels of abstraction provided by the RISC-V dialect down to regular 
+assembly should not be difficult.
 
@@
 
-Again, the core contribution will be proof automation. By treating
+The core contribution will be proof automation. By treating
 instruction selection as a sequence of rewrites, we merely need to
@@
 investigate ways to simplify the model under specific assumptions. For
-example, by assuming we're only compiling user-level programs, we might
-be able to simplify away any complexity to do with kernel-level
+example, we could assume we're only compiling user-level programs, which might
+allow us to simplify away any complexity to do with kernel-level
 instructions in the model.
@@
 simulator as a basis for the translation validation tool I aim to
-develop, adapting it to the slightly different design constraints in
+develop in this project, adapting it to the slightly different design constraints in
 this context.
@@
 theorem prover~\cite{sammlerIslarisVerificationMachine2022}. Their strategy starts by using Isla to transform a program into a trace, and then proving properties about this trace.
-To remove the need to trust the SMT solver from, Sammler et al also investigated
+To remove the need to trust the SMT solver, Sammler\etal{} also investigated
 the use of automated translation validation, which certifies that the trace produced
@@
 worked well for RISC-V, but unfortunately did not scale to the much larger Arm model.
+
+Although I'll focus specifically on RISC-V, I'll still choose to reimplement 
+the symbolic simulator in Lean, rather than reusing Isla in the same way, 
+in the hopes that this approach might prove more scalable.
 
@@
 and Lean does not support coinductive types.
-I'd like to address this, and build an ergonomic Lean library for ITrees,
+This is why I'll initially be building a semantics for a fragment of the IR with
+trivially terminating for-loops. 
+In the current chapter I'll address this shortcoming, by building a Lean library for ITrees and using it to define semantics for potentially infinite loops,
 answering research question \textsc{RQ3} in the process.
@@
 Luckily, we can encode coinductives in Lean's existing logic, using
-so-called \emph{quotients of polynomial functors} (QPFs). Avigad et
-al.~have written about the low-level construction~\cite{avigadDataTypesQuotients2019}, 
+so-called \emph{quotients of polynomial functors} (QPFs). Avigad\etal{} have written about the low-level construction~\cite{avigadDataTypesQuotients2019}, 
 upon which I expanded in my Master's thesis by prototyping a higher-level framework for defining
@@
 the low-level corecursion principle to work with. However, in their
-original development of ITrees, Xia et al. in fact provide a set of three
+original development of ITrees, Xia\etal{} in fact provide a set of three
 combinators which they claim make working with ITrees more tractable,
@@
 in Lean, we don't have to worry about general corecursive functions; it
-suffices to implement the three basic combinators that Xia et
-al.~provide.
+suffices to implement the three basic combinators that Xia\etal{} provide.
 
@@
 
-Isabelle, in particular, is relevant because it's construction of
+Isabelle, in particular, is relevant because its construction of
 coinductive types, in terms of \emph{bounded natural functors} is
@@
 
+As mentioned, I'll work on the instruction selection project last.
+I'll list the 3 chapters in order that I'll work on them, which differs from the
+chapter order presented so far.
+
 \subsection*{High-Assurance Translation Validation for Compiler Optimizations}
@@
 
-\subsection*{Scaling High-Assurance Translation Validation to
-Authoritative ISA
-Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models-1}
+\subsection*{Automatically Verified Instruction Selection using
+Authoritative ISA Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models-1}
 

### 2025-01-31 paper.tex
remove dot from title
add progress report & conclusion

@@
 for Declarative
-Transformations.}                                      
+Transformations}
 \subtitle{First Year Report}
@@
 semantics we need without referring to details of the underlying QPF construction directly.
-
-\subsubsection*{Verified Compilation and Lean's Extensibility}
-
-Note that the benefits of a verified compilation pipeline go far beyond
-simply reducing the chance of bugs. It could lead to a culture
-shift in the way software is written. Traditionally, writing compiler
-optimizations was a domain reserved for the wise elders who we could
-trust to perform the appropriate ``black magic''. Reasoning about the
-correctness of a transformation was simply too hard, so we had to rely
-on interpersonal trust. Tools like Alive have already caused a shift in
-this approach: nowadays PRs that propose adding a new peephole
-optimization to LLVM can include a link to Alive's online interface
-showing the correctness, meaning no further discussion or review is
-required. However, this effect is still limited to making reviews for
-proposed additions to LLVM easier.
-
-Lean, again, presents a very interesting opportunity because of it's
-\emph{extensibilty}: Lean already has an extensible syntax and a
-powerful meta-programming API. In fact, a lot of features that are
-considered part of Lean's core are actually implemented using the same
-meta-programming facilities that are available to libraries (we'll use
-this extensively in~\autoref{a-codatatypes-library-for-lean}). However, we imagine a future where this
-extensibility could extend to the backend of the compiler also, such
-that a library writer might implement an algorithm in a functional style
-that is easy to reason about, but also provide an optimized assembly
-\emph{and} a proof that the assembly is equivalent. Traditionally,
-hard-coded assembly in a library has been cause for increased scrutiny.
-Optimizing is hard, do we really trust the author of this library to
-have done it correctly? Verification means we no longer have to trust,
-instead we get to have fearless bespoke optimizations!
 
@@
 
+\section{Progress Report}\label{progress-report}
+
+So far, I've worked on laying the foundations for my proposed research, in the form of the Lean-MLIR framework and bitvector decision procedures in Lean.
+
+\subsection{Lean-MLIR}\label{lean-mlir}
+
+As mentioned before, Lean-MLIR is a framework for giving semantics to MLIR dialects in Lean.
+On top of this framework, we've built a peephole rewriter, and proved
+that if a rewrite locally preserves semantics, then applying this
+rewrite in the context of a larger program will preserve the semantics
+of the program as a whole.
+
+However, we've focused on dialects with pure (i.e., side effect free)
+semantics. Even though we \emph{are} able to verify a lot of interesting
+rewrites in high-level MLIR dialects, we do need to be able to reason
+about side effect if we want to capture the low-level dialects as well.
+
+To this end, I've\footnote{The framework is a collaborative effort; the handling of side effect is one of my contributions.} generalized the framework so that a dialects
+semantics may be given in a user-defined monad, and proven that applying
+a semantics-preserving pure rewrite (i.e., where the operations being
+rewritten are all side effect free) inside a side-effecting program
+preserves semantics of the program as a whole.
+
+\subsection{Bitvectors}\label{bitvectors}
+
+As mentioned, a big reason for us to use Lean is its community. As
+such, I've been heavily involved in the redesign of Lean's API for
+bitvectors, and the move of bitvectors from Mathlib to become part of Lean's core library
+(i.e., the library that is by default included in every Lean program).
+By ensuring our representation of bitvectors is aligned with the rest of
+the community, we ensure we can easily reap the benefits of, e.g.,
+tactics to automate bitvector proofs made by others.
+
+One such effort is the verified bitblaster and SAT solver integration written by the Lean FRO.
+Siddharth (another PhD student in our group) and I have contributed to this project by 
+formally proving equivalences between bitvector operations and bitblast-able circuits.
+
+As the semantics of both LLVM IR and assembly heavily involves bitvectors, much of 
+the automation I propose will be built on top of this bitvector decision procedure.
+
+\subsection{Amazon Internship}\label{amazon-internship}
+
+At Amazon, I wrote a scalable symbolic simulator for straight-line, memory-free Arm machine code, in the context of LNSym, 
+a toolkit for the verification of cryptographic hash functions and block ciphers written in Arm assembly.
+Before my internship, a prototype symbolic simulator existed, but it was limited in features and did not scale well.
+My reimplemented version tracks more information of the simulated execution, 
+and is capable of simulating more than 400 instructions of the SHA512 litmus test we benchmarked against before hitting Lean's default time-out limits.
+This is about 10x more instructions than the prototype was able to simulate given the same time-out limits.
+
+The prototype relied heavily on Lean's builtin simplifier. 
+I replaced this with bespoke metacode, that was able to exploit the specific shape of the problem we were solving.
+Firstly, I wrote preprocessing that allowed us to treat the operational semantic model, where operations are given semantics as functions that write to state, 
+in an axiomatic-semantics way, by having theorems that characterizes operations in terms of reads from the resulting state.
+Secondly, I wrote a bespoke data-structure and used it selectively reflects parts of the proof goal.
+We explicitly decoupled this meta-code needed to scale symbolic simulation from the definition of the semantic model,
+so that the model remains extensible even by people who are not experts in formal verification, which was an explicit design goal.
+
+This project, like some of my other research, was also an experiment in the usability of Lean as a platform to verify software.
+In particular, we generated very large proofs, and exposed, amongst others, a fundamental scalability problem in Lean:
+The time Lean needs to assemble a proof after a tactic executes scales roughly by the length of the proof times the number of binders in the proof.
+In the larger benchmarks, LNSym would generate proofs involving tens of thousands of binders,
+resulting in the majority of time needed to check the proof was actually spent in this proof-assembling step.
+We reported this to the Lean core team, who are actively working on a solution.
+Furthermore, the semantic model of Arm machine code used relies heavily on bitvectors, thus, we also used the proof obligations resulting from the improved symbolic simulator to stress-test the bitvector decision procedure.
+I've reported these problems, and submitted patches for some, contributing to the greater vision of improving the usability of Lean.
+
+\section{Conclusion}
+
+Implementation a verified compilation pipeline as described in this proposal could 
+have an impact far beyond simply reducing the chance of bugs. It could lead to a culture
+shift in the way software is written. Traditionally, writing compiler
+optimizations was a domain reserved for the wise elders who we could
+trust to perform the appropriate ``black magic''. Reasoning about the
+correctness of a transformation was simply too hard, so we had to rely
+on interpersonal trust. Tools like Alive have already caused a shift in
+this approach: nowadays PRs that propose adding a new peephole
+optimization to LLVM can include a link to Alive's online interface
+showing the correctness, meaning no further discussion or review is
+required. However, this effect is still limited to making reviews for
+proposed additions to LLVM easier.
+
+Lean, again, presents a very interesting opportunity because of it's
+\emph{extensibilty}: Lean already has an extensible syntax and a
+powerful meta-programming API. In fact, a lot of features that are
+considered part of Lean's core are actually implemented using the same
+meta-programming facilities that are available to libraries (we'll use
+this extensively in~\autoref{a-codatatypes-library-for-lean}). However, we imagine a future where this
+extensibility could extend to the backend of the compiler also, such
+that a library writer might implement an algorithm in a functional style
+that is easy to reason about, but also provide an optimized assembly
+\emph{and} a proof that the assembly is equivalent. Traditionally,
+hard-coded assembly in a library has been cause for increased scrutiny.
+Optimizing is hard, do we really trust the author of this library to
+have done it correctly? Verification means we no longer have to trust,
+instead we get to have fearless bespoke optimizations!
+
 %% Bibliography

### 2025-01-31 paper.tex
Tobias' comments

@@
 develop proof automation to show local
-correctness of a transformation, without needing input from a formal methods export?
+correctness of a transformation, without needing a formal methods expert?
 
@@
 \cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
-included no proof automation. This will, to my knowledge, be the first
+did not provide automated optimization verification. This will, to my knowledge, be the first
 effort to provide fully automated proofs for correctness of LLVM
@@
 Finally, to model the semantics of LLVM IR in a modular way, we'd like
-to follow Zakowski et
-al.~(\cite{zakowskiModularCompositionalExecutable2021}) and use
-interaction trees (\cite{xiaInteractionTreesRepresenting2020}).
+to follow Zakowski\etal{}~\cite{zakowskiModularCompositionalExecutable2021} and use
+interaction trees~\cite{xiaInteractionTreesRepresenting2020}.
 Unfortunately, interaction trees are \emph{coinductive} in nature, to
@@
 
-My research will aim to answer the following questions:
+My research aims to answer the following questions:
 \begin{enumerate}
@@
 IR that contains bitvector arithmetic (a fragment of the \texttt{arith}
-dialect) memory manipulations (\texttt{ptr} dialect) and structured
+dialect), memory manipulations (\texttt{ptr} dialect) and structured
 control flow (i.e., for and while loops, \texttt{scf} dialect). 
@@
 peephole rewriting as a way to lower proof burden in CompCert
-development, but they still prove these rewrites manually.
+development, but they still require manual effort to prove these rewrites.
 
-Katamaran \cite{keuchelVerifiedSymbolicExecution2022} is a
+Katamaran~\cite{keuchelVerifiedSymbolicExecution2022} is a
 symbolic simulation tool written in Coq. However, this
@@
 
-Implementation a verified compilation pipeline as described in this proposal could 
+Implementing a verified compilation pipeline as described in this proposal could 
 have an impact far beyond simply reducing the chance of bugs. It could lead to a culture
